\documentclass[11pt, twoside]{report}

\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage[bitstream-charter]{mathdesign}
\usepackage{bbding}
\usepackage{ragged2e}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{paracol}
\usepackage{mdframed}
\usepackage[margin=1in]{geometry}

\usepackage[autocompile]{gregoriotex}

\titleformat{\chapter}[block]{\huge\scshape\filcenter}{}{1em}{}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{}{1em}{}

\mdfsetup{skipabove=\topskip, skipbelow=\topskip}

\newcommand{\rubric}[1]{
	\switchcolumn[0] {
		\itshape
		#1
	}
}

\newcommand{\latinenglish}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape\small
		#2
	}
}

\newcommand{\latinenglishequal}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape
		#2
	}
}

\newenvironment{latinenglishsection}
	{\columnratio{.7, .3} \begin{paracol}{2}}
	{\end{paracol}}

\newenvironment{latinenglishequalsection}
	{\columnratio{.5, .5}\begin{paracol}{2}}
	{\end{paracol}}

\setlength{\columnseprule}{0.4pt}

\newcommand{\heading}[1]{
	\begin{leftcolumn}
		#1
	\end{leftcolumn}
}

\newcommand{\spanning}[1]{
	\switchcolumn*[#1]
}

\newenvironment{verses}[1]
	{\begin{flushleft} \begin{enumerate}[leftmargin=*] \setcounter{enumi}{#1}}
	{\end{enumerate} \end{flushleft}}

\newenvironment{versicles}{\par\leavevmode\parskip=0pt}{}

\newenvironment{collect}
{
	\leavevmode
	\parindent=1em
	\parskip=0pt
	\noindent Orémus.\par
}{}

\newenvironment{optionbox}
{
	\switchcolumn[0]
	\begin{mdframed}
%	\begin{minipage}{0.8\linewidth}
}{
%	\end{minipage}
	\end{mdframed}
}

\newcommand{\optionrule}{
	\begin{center}
	\rule{0.5\linewidth}{0.6pt}
	\end{center}
}

\newenvironment{optionruled}
{
	\optionrule
}
{
	\optionrule
}

% for use inside the collect environment
\newcommand{\Amen}{\par\noindent \Rbar. Amen.}

\begin{document}

\title{Vespers of Our Lady\\In Advent}
\date{ }

\maketitle

PLACEHOLDER FRONT MATTER

\chapter*{Vespers}

\textit{This form of the Office of Vespers is celebrated for the first time on the Saturday before the first Sunday of Advent,
and for the last time on December 23rd, that is, the evening before Christmas Eve.
It is also celebrated on the Feast of the Annunciation.
The Office of the Annunciation, as a result of falling either in Lent or in Eastertide,
requires some variations that are not shown in this booklet.}

\section*{Preparatory Prayers}

\textit{ All kneel and pray silently. As you say the prayer \textnormal{Aperi, Domine}, make the sign of the cross with your thumb first over your lips, and then over your heart.}

\begin{latinenglishequalsection}

\latinenglishequal{
	Áperi, \CrossMaltese\ Dómine, os meum ad benedicéndum\linebreak nomen sanctum tuum:
	\CrossMaltese\ munda quoque cor meum ab ómnibus vanis, pervérsis et aliénis cogitatiónibus;
	intelléctum illúmina, afféctum inflámma, ut digne, atténte ac devóte hoc Offícium beátæ Vírginis Maríæ recitáre váleam,
	et exaudíri mérear ante conspéctum divínæ Majestátis tuæ.
	Per Christum Dóminum nostrum. 
	Amen.
}{
	Open, + O Lord, my mouth to bless Thy holy Name; + cleanse also my heart from all vain, evil, and wandering thoughts; enlighten my understanding and kindle my affections; that I may worthily, attentively, and devoutly say this Office of the Blessed Virgin Mary, and so merit to be heard before the presence of Thy divine Majesty.  Through Christ our Lord.  Amen.
}

\latinenglishequal{
	Domine, in unióne illíus divínæ intentiónis, qua ipse in terris laudes Deo persolvísti, has tibi Horas persólvo.
}{
	O Lord, in union with that divine intention wherewith thou, whilst here on earth, didst render praises unto God, I desire to offer this my Office of prayer unto thee.
}

\latinenglishequal{
	Ave María, grátia plena, Dóminus tecum. Benedíc\-ta tu in muliéribus, et benedíctus fructus ventris tui, Jesus.
 Sancta María, Mater Dei, ora pro nobis peccatóribus, nunc et in hora mortis nostræ. Amen.
 }{
 	Hail Mary, full of grace, the Lord is with thee. Blessed art thou among women, and blessed is the fruit of thy womb, Jesus.
 Holy Mary, Mother of God, pray for us sinners, now and at the hour of our death. Amen.
 }
 
 \end{latinenglishequalsection}
 \pagebreak
 
 \begin{latinenglishsection}
 
\heading{\section*{Invitatory}}

\rubric{As the officiant intones the beginning of the office, all stand and make the sign of the cross.}

\latinenglish{
	\gregorioscore{deus-in-adjutorium}
}{
	O God, come to my assistance.
		
	O Lord, make haste to help me.
	
	Glory be to the Father, and to the Son, and to the Holy Spirit,
	as it was in the beginning, is now, and ever shall be, world without end. Amen.
	
	Alleluia.
}

\heading{\section*{Psalm 109}}

\rubric{The officiant intones the antiphon:}

\latinenglish{
	\textbf{Ant. 1.} Missus est.
}{
	The angel.
}

\rubric{Then the cantor sings the first half of the first verse of the psalm, and the choir continues, alternating sides between each verse of the psalm.}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm1-advent}
	
	\begin{verses}{1}

	\item Donec ponam inimícos \textbf{tu}os,~*\linebreak
	scabéllum pe\textit{dum} \textit{tu}\textbf{ó}rum.

	\item Virgam virtútis tuæ emíttet Dóminus ex \textbf{Si}on:~*\linebreak
	domináre in médio inimicó\textit{rum} \textit{tu}\textbf{ó}rum.
	
	\item Tecum princípium in die virtútis tuæ in splendóribus sanc\textbf{tó}rum:~*\linebreak
	ex útero ante lucíferum \textit{gé}\textit{nu}\textbf{i} te.
	
	\item Jurávit Dóminus, et non pœnitébit \textbf{e}um:~*\linebreak
	Tu es sacérdos in ætérnum secúndum órdi\textit{nem} \textit{Mel}\textbf{chí}sedech.
	
	\item Dóminus a dextris \textbf{tu}is,~*\linebreak
	confrégit in die iræ \textit{su}\textit{æ} \textbf{re}ges.
	
	\item Judicábit in natiónibus, implébit ru\textbf{í}nas:~*\linebreak
	conquassábit cápita in ter\textit{ra} \textit{mul}\textbf{tó}rum.
	
	\item De torrénte in via \textbf{bi}bet:~*\linebreak
	\textit{(stand)} proptérea exal\textit{tá}\textit{bit} \textbf{ca}put.
	
	\item \textit{(bow)} Glória Patri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item \textit{(rise)} Sicut erat in princípio, et nunc, et \textbf{sem}per,~*\linebreak
	\textit{(sit)} et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. The Lord said to my Lord:
	Sit thou at my right hand,
	
	2. Until I make thine enemies:
	thy footstool.
	
	3. The Lord shall send forth the rod of thy power from out of Sion:
	rule Thou in the midst of thine enemies.
	
	4. Thine shall be the dominion in the day of thy power, amid the brightness of the saints:
	from the womb, before the day-star, have I begotten thee.
	
	5. The Lord hath sworn, and will not repent:
	Thou art a priest forever according to the order of\linebreak Melchisedech.
	
	\pagebreak
	
	6. The Lord upon thy right hand:
	hath overthrown kings in the day of his wrath.
	
	7. He shall judge among the nations, he shall fulfil destructions:
	he shall smite in sunder the heads in the land of many.
	
	8. He shall drink of the brook in the way:
	therefore shall he lift up his head.
	
	Glory be.
}
	
\rubric{The same postures are observed at the end of every psalm.}
	
\rubric{All sing the full antiphon together:}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant1-advent}
}{
	\textnormal{Ant.}
	The angel Gabriel was sent to Mary, a virgin espoused to Joseph.
}

\rubric{The remaining psalms with their antiphons are sung in the same way as the first.}

\heading{\section*{Psalm 112}}

\latinenglish{
	\textbf{Ant. 2.} Ave María.
}{
	Hail Mary.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm2-advent}
	
	\begin{verses}{1}
	
	\item \textit{(bow head)} Sit nomen Dómini \textbf{be}ne\textbf{díc}tum,~*\linebreak
	ex hoc nunc, et us\textit{que} \textit{in} \textbf{s\'{\ae}}culum.

	\item A solis ortu usque \textbf{ad} oc\textbf{cá}sum,~*\linebreak
	laudábile \textit{no}\textit{men} \textbf{Dó}mini.
	
	\item Excélsus super omnes \textbf{gen}tes \textbf{Dó}minus,~*\linebreak
	et super cælos gló\textit{ri}\textit{a} \textbf{e}jus.
	
	\item Quis sicut Dóminus, Deus noster, qui in \textbf{al}tis \textbf{há}bitat,~*\linebreak
	et humília réspicit in cælo \textit{et} \textit{in} \textbf{ter}ra?
	
	\item Súscitans a \textbf{ter}ra \textbf{ín}opem,~*\linebreak
	et de stércore é\textit{ri}\textit{gens} \textbf{páu}perem:
	
	\item Ut cóllocet eum \textbf{cum} prin\textbf{cí}pibus,~*\linebreak
	cum princípibus pó\textit{pu}\textit{li} \textbf{su}i.
	
	\pagebreak
	
	\item Qui habitáre facit stéri\textbf{lem} in \textbf{do}mo,~*\linebreak
	matrem filió\textit{rum} \textit{læ}\textbf{tán}tem.
	
	\item Glória \textbf{Pa}tri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.

	\end{verses}
}{
	 1.  Praise the Lord, ye children:\linebreak praise ye the name of the Lord.
	 
	 2. Blessed be the name of the Lord, from henceforth now and for ever.
	 
	 3. From the rising of the sun unto the going down of the same, the name of the Lord is worthy of praise.
	 
	 4. The Lord is high above all nations; and his glory above the heavens.
	 
	 5. Who is as the Lord our God, who dwelleth on high: 
	 And looketh down on the low things in heaven and in earth?
	
	6. Raising up the needy from the earth, and lifting up the poor out of the dunghill:
	
	7. That he may place him with princes, with the princes of his people.
	
	8. Who maketh a barren woman to dwell in a house, the joyful\linebreak mother of children.
	
	Glory be.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant2-advent}
}{
	\textnormal{Ant.}
	Hail Mary, full of grace, the Lord is with thee: blessed art thou among women, alleluia.
}

\heading{\section*{Psalm 121}}

\latinenglish{
	\textbf{Ant. 3.} Ne tímeas.
}{
	Fear not.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm3-advent}

	\begin{verses}{1}
	
	\item Stantes erant pedes \textbf{nos}tri,~*\linebreak
	in átriis tu\textit{is}, \textit{Je}\textbf{rú}salem.

	\item Jerúsalem, quæ ædificátur ut \textbf{cí}vitas:~*\linebreak
	cujus participátio ejus \textit{in} \textit{id}\textbf{íp}sum.
	
	\item Illuc enim ascendérunt tribus, tribus \textbf{Dó}mini:~* \linebreak
	estimónium Israël ad confiténdum nó\textit{mi}\textit{ni} \textbf{Dó}mini.
	
	\item Quia illic sedérunt sedes in ju\textbf{dí}cio,~*\linebreak
	sedes super \textit{do}\textit{mum} \textbf{Da}vid.
	
	\item Rogáte quæ ad pacem sunt Je\textbf{rú}salem:~*\linebreak
	et abundántia dili\textit{gén}\textit{ti}\textbf{bus} te:
	
	\item Fiat pax in virtúte \textbf{tu}a:~*\linebreak
	et abundántia in túr\textit{ri}\textit{bus} \textbf{tu}is.
	
	\item Propter fratres meos, et próximos \textbf{me}os,~*\linebreak
	loquébar \textit{pa}\textit{cem} \textbf{de} te:
	
	\item Propter domum Dómini, Dei \textbf{nos}tri,~*\linebreak
	quæsívi \textit{bo}\textit{na} \textbf{ti}bi.
	
	\item Glória Patri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et nunc, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. I rejoiced at the things that were said to me: We shall go into the house of the Lord.
	
	2. Our feet were standing in thy courts, O Jerusalem.
	
	3. Jerusalem, which is built as a city, which is compact together.
	
	4. For thither did the tribes go up, the tribes of the Lord: the testimony of Israel, to praise the name of the Lord.
	
	5. Because their seats have sat in judgment, seats upon the house of David.

	6. Pray ye for the things that are for the peace of Jerusalem: and\linebreak abundance for them that love thee.
	
	7. Let peace be in thy strength: and abundance in thy towers.
	
	8. For the sake of my brethren, and of my neighbours, I spoke\linebreak peace of thee.
	
	9. Because of the house of the Lord our God, I have sought good things for thee.
	
	\pagebreak
	
	Glory be. 
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant3-advent}
}{
	\textnormal{Ant.}
	Fear not, Mary, thou hast found grace with the Lord: behold, thou shalt conceive, and bear a son, alleluia.
}

\heading{\section*{Psalm 126}}

\latinenglish{
	\textbf{Ant. 4.} Dabit ei Dóminus.
}{
	The Lord God shall give.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm4-advent}
	
	\begin{verses}{1}
	
	\item Nisi Dóminus custodíerit \textbf{ci}vi\textbf{tá}tem,~*\linebreak
	frustra vígilat qui cus\textit{tó}\textit{dit} \textbf{e}am.

	\item Vanum est vobis ante \textbf{lu}cem \textbf{súr}gere:~*\linebreak
	súrgite postquam sedéritis, qui manducátis pa\textit{nem} \textit{do}\textbf{ló}ris.
	
	\item Cum déderit diléctis \textbf{su}is \textbf{som}num:~*\linebreak
	ecce heréditas Dómini fílii: merces, \textit{fruc}\textit{tus} \textbf{ven}tris.
	
	\item Sicut sagíttæ in \textbf{ma}nu pot\textbf{én}tis:~*\linebreak
	ita fílii \textit{ex}\textit{cus}\textbf{só}rum.
	
	\item Beátus vir qui implévit desidérium \textbf{su}um ex \textbf{ip}sis:~*\linebreak
	non confundétur cum loquétur inimícis su\textit{is} \textit{in} \textbf{por}ta.
	
	\item Glória \textbf{Pa}tri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. Unless the Lord build the house, they labour in vain that build it.
	
	2. Unless the Lord keep the city, he watcheth in vain that keepeth it.
	
	3. It is vain for you to rise before light, rise ye after you have rested, you that eat the bread of sorrow.
	
	4. When he shall give sleep to his beloved: behold the inheritance of the Lord are children: the reward, the fruit of the womb.
	
	5. As arrows in the hand of the mighty, so the children of them that have been shaken.
	
	6. Blessed is the man that hath filled the desire with them;
	he shall not be confounded when he shall speak to his enemies in the gate. 
	
	Glory be.
}

\flushpage

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant4-advent}
}{
	\textnormal{Ant.}
	The Lord God shall give unto him the throne of David his father: and he shall reign forever.
}

\heading{\section*{Psalm 147}}

\latinenglish{
	\textbf{Ant. 5.} Ecce ancílla Dómini.
}{
	Behold the handmaid of the Lord.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm5-advent}
	
	\begin{verses}{1}
	
	\item Quóniam confortávit seras portárum tu\textbf{á}rum:~*\linebreak
	benedíxit fíliis \textit{tu}\textit{is} \textbf{in} te.

	\item Qui pósuit fines tuos \textbf{pa}cem:~*\linebreak
	et ádipe fruménti \textit{sá}\textit{ti}\textbf{at} te.
	
	\item Qui emíttit elóquium suum \textbf{ter}ræ:~*\linebreak
	velóciter currit \textit{ser}\textit{mo} \textbf{e}jus.
	
	\item Qui dat nivem sicut \textbf{la}nam:~*\linebreak
	nébulam sicut cí\textit{ne}\textit{rem} \textbf{spar}git.
	
	\item Mittit crystállum suam sicut buc\textbf{cél}las:~*\linebreak
	ante fáciem frígoris ejus quis \textit{sus}\textit{ti}\textbf{né}bit?
	
	\item Emíttet verbum suum, et liquefáciet \textbf{e}a:~*\linebreak
	flabit spíritus ejus, et \textit{flu}\textit{ent} \textbf{a}quæ.
	
	\item Qui annúntiat verbum suum \textbf{Ja}cob:~*\linebreak
	justítias, et judícia \textit{su}\textit{a} \textbf{Is}raël.
	
	\item Non fecit táliter omni nati\textbf{ó}ni:~*\linebreak
	et judícia sua non manifes\textit{tá}\textit{vit} \textbf{e}is.
	
	\item Glória Patri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et nunc, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	 1.  Praise the Lord, O Jerusalem: praise thy God, O Sion. 
	 
	 2. Because he hath strengthened the bolts of thy gates, he hath\linebreak blessed thy children within thee. 
	 
	 3. Who hath placed peace in thy borders: and filleth thee with the fat of corn.
	 
	 4. Who sendeth forth his speech to the earth: his word runneth swiftly.
	 
	 5. Who giveth snow like wool: scattereth mists like ashes.
	
	6. He sendeth his crystal like\linebreak morsels: who shall stand before the face of his cold?
	
	7. He shall send out his word, and shall melt them: his wind shall\linebreak blow, and the waters shall run.
	
	8. Who declareth his word to Jacob: his justices and his judgments to Israel.
	
	9. He hath not done in like manner to every nation: and his judgments he hath not made manifest to them.
	
	Glory be.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant5-advent}
}{
	\textnormal{Ant.}
	Behold the handmaid of the Lord: be it done unto me according to thy word.
}

\heading{\section*{Little Chapter (Isaiah 11:1-2)}}

\latinenglish{
	\textit{All stand.}
	
	Egrediétur virga de radíce Jesse,~\GreDagger\
	et flos de radíce ejus ascéndet,~*
	et requiéscet super eum Spíritus Dómini.
	
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) <sp>R/</sp>. De(h)o(h) grá(f)ti(e)as.(e.f.) (::)
	}
}{
	And there shall come forth a rod out of the root of Jesse,
	and a flower shall rise up out of his root,
	and the spirit of the Lord shall rest upon him.
	\Rbar.~Thanks be to God.
}

\heading{\section*{Hymn}}

\rubric{All kneel during the first verse of this hymn.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ave-maris-stella}
	
	\begin{verses}{1}
	
	\item \textit{(stand)} Sumens illud Ave, Gabriélis ore,\linebreak
	funda nos in pace, mutans Hevæ nomen.
	
	\item Solve vincla reis, profer lumen cæcis,\linebreak
	mala nostra pelle, bona cuncta posce.
	
	\item Monstra t\textit{e} esse matrem, sumat per te preces,\linebreak
	qui pro nobis natus, tulit esse tuus.
	
	\item Virgo singuláris, inter omnes mitis,\linebreak
	nos culpis solútos, mites fac et castos.
	
	\item Vitam præsta puram, iter para tutum,\linebreak
	ut vidéntes Jesum semper collætémur.
	
	\item \textit{(bow)} Sit laus Deo Patri, summo Christo decus,\linebreak
	Spirítui Sancto, tribus honor unus. Amen.
	
	\end{verses}
}{
	1.  Hail, O Star of the ocean,
	God's own Mother blest,
	ever sinless Virgin,
	gate of heav'nly rest. 
	
	2. Taking that sweet Ave,
	which from Gabriel came,
	peace confirm within us,
	changing Eve's name.
	
	3. Break the sinners' fetters,
	make our blindness day,
	chase all evils from us,
	for all blessings pray. 
	
	4. Show thyself a Mother,
	may the Word divine,
	born for us thine\linebreak Infant,
	hear our prayers through\linebreak thine.
	
	5. Virgin all excelling,
	mildest of the mild,
	free from guilt preserve us
	meek and undefiled.
	
	6. Keep our life all spotless,
	make our way secure,
	till we find in Jesus
	joy for evermore. 
	
	7. Praise to God the Father,
	honor to the Son,
	in the Holy Spirit,
	be the glory one. Amen. 
}

\flushpage

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{vers-diffusa-est}
}{
	Grace was poured forth on thy lips.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{vers-propterea-benedixit}
}{
	Therefore hath the Lord blessed thee for ever.
}

\heading{\section*{Canticle of Our Lady (Luke 1:46-55)}}

\rubric{Remain standing throughout the canticle.}

\latinenglish{
	\textbf{Ant. Magn.} Spíritus Sanctus.
}{
	The Holy Spirit.
}

\rubric{As the cantor intones the \textnormal{Magnificat}, make the sign of the cross.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{magnificat-advent}
	
	\begin{verses}{2}
	
	\item Quia respéxit humilitátem \textit{an}\textit{cíl}\textit{læ} \textbf{su}æ:~*\linebreak
	ecce enim ex hoc beátam me dicent omnes gene\textit{ra}\textit{ti}\textbf{ó}nes.
	
	\item Quia fecit mihi \textit{ma}\textit{gna} \textit{qui} \textbf{pot}ens est:~*\linebreak
	et sanctum \textit{no}\textit{men} \textbf{e}jus.
	
	\item Et misericórdia ejus a progéni\textit{e} \textit{in} \textit{pro}\textbf{gé}nies~*\linebreak
	timén\textit{ti}\textit{bus} \textbf{e}um.
	
	\item Fecit poténtiam in \textit{brá}\textit{chi}\textit{o} \textbf{su}o:~*\linebreak
	dispérsit supérbos mente \textit{cor}\textit{dis} \textbf{su}i.
	
	\item Depósuit pot\textit{én}\textit{tes} \textit{de} \textbf{se}de,~*\linebreak
	et exal\textit{tá}\textit{vit} \textbf{hú}miles.
	
	\item Esuriéntes \textit{im}\textit{plé}\textit{vit} \textbf{bo}nis:~*\linebreak
	et dívites dimí\textit{sit} \textit{in}\textbf{á}nes.
	
	\item Suscépit Israël \textit{pú}\textit{e}\textit{rum} \textbf{su}um,~*\linebreak
	recordátus misericór\textit{di}\textit{æ} \textbf{su}æ.
	
	\item Sicut locútus est \textit{ad} \textit{pa}\textit{tres} \textbf{nos}tros,~*\linebreak
	Abraham et sémini e\textit{jus} \textit{in}\textbf{s\'{\ae}}cula.
	
	\item \textit{(bow)} Glória \textit{Pa}\textit{tri}, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item \textit{(rise)} Sicut erat in princípio, \textit{et} \textit{nunc}, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. My soul doth magnify the Lord.
	
	2.  And my spirit hath rejoiced in God my Saviour.
	
	3. Because he hath regarded the humility of his handmaid;
	for\linebreak behold from henceforth all\linebreak generations shall call me blessed.
	
	4. Because he that is mighty, hath done great things to me; and holy is his name. 
	
	5. And his mercy is from generation unto generations, to them that fear him. 

	6. He hath shewed might in his arm: he hath scattered the proud in the conceit of their heart.
	
	7. He hath put down the mighty from their seat, and hath exalted the humble.
	
	8. He hath filled the hungry with good things; and the rich he hath sent empty away.
	
	9. He hath received Israel his servant, being mindful of his mercy:
	
	10. As he spoke to our fathers, to Abraham and to his seed for ever.
	
	Glory be.
}

\rubric{All sit.}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{antcant-advent}
}{
	\textnormal{Ant.}
	The Holy Spirit shall come upon thee, Mary: fear not, thou shalt bear in thy womb the Son of God, alleluia.
}


\heading{\section*{Collect}}

\rubric{All stand.}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{	
	\begin{collect}
	Deus, qui de beátæ Maríæ Vírginis utero, Verbum tuum, ángelo nuntiánte,
	carnem suscípere voluísti; præsta supplícibus tuis, ut qui vere eam Genitrícem Dei crédimus,
	ejus apud te intercessiónibus adjuvémur.
	Per eumdem Dominum nostrum.
	\Amen
	\end{collect}
}{
	Let us pray.
	O God, who wast\linebreak pleased that Thy Word, at the message of an angel,
	should take flesh in the womb of the blessed Virgin Mary;
	grant to Thy humble serv\-ants, that we who believe her to be truly the Mother of God,
	may be assisted by her intercessions with Thee.
	\Rbar.~Amen.
}

\rubric{Commemorations or votive suffrages may be said here.}

\flushpage

\heading{\section*{Suffrage of St. Joseph}}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant-joseph}
	
	\begin{versicles}
	\Vbar. Ecce homo sine queréla, verus Dei cultor.
	
	\Rbar. Ábstinens se ab omni ópere malo, et pérmanens in innocéntia sua.
	\end{versicles}
	
	\begin{collect}
	Sanctíssimæ Genetrícis tuæ Sponsi, qu\'{\ae}sumus, Dómine, méritis adjuvémur: ut, quod possibílitas nostra non óbtinet, ejus nobis intercessióne donétur.
	%(Qui vivis.)
	\end{collect}
}{
	A faithful and wise servant, whom the Lord hath set over His household.
	
	Behold a man without blame, a true worshipper of God.
	\Rbar.~Abstaining from every evil work, and abiding in his innocence.
	
	Let us pray.
	Assist us, O Lord, we beseech thee,
	by the merits of the spouse of thy most holy Mother,\linebreak
	that what of ourselves we are unable to obtain,
	may be granted to us by his intercession.
}

\heading{\section*{Suffrage of the Saints}}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant-all-saints-advent}
	
	\begin{versicles}
	\Vbar. Ecce, apparébit Dóminus super nubem cándidam.
	
	\Rbar. Et cum eo Sanctórum míllia.
	\end{versicles}
	
	\begin{collect}
	Consciéntias nostras, qu\'{\ae}sumus, Dómine, visitándo purífica;
	ut vén\-iens Jesus Christus Fílius tuus Dóminus noster, cum ómnibus Sanctis,
	parátam sibi in nobis invéniat mansiónem.
	Qui tecum vivit et regnat.
	\Amen
	\end{collect}
}{
	Behold, the Lord shall come, and all his Saints with him,
	and in that day there shall be a great light, alleluia.
	
	Behold, the Lord shall appear upon a shining cloud.
	\Rbar.~And with him thousands of Saints.
	
	Let us pray.
	Visit, O Lord, we\linebreak beseech Thee, and purify our consciences;
	that Jesus Christ Thy Son our Lord,
	when He cometh with all His Saints,
	may find in us a mansion prepared for him.
	Who liveth and reigneth.
	 \Rbar.~Amen.
}

%\flushpage

\heading{\section*{Conclusion}}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{benedicamus-domino}
}{
	Let us bless the Lord.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{deo-gratias}
}{
	Thanks be to God.
}

%\begin{optionbox}
%
%	\rubric{Or, if this day or the following is a Sunday of Advent:}
%	\gresetinitiallines{0}
%	\gregorioscore{benedicamus-advent}
%	\gregorioscore{deo-gratias-advent}
%
%\end{optionbox}

\latinenglish{
	\begin{versicles}
	\Vbar. Fidélium ánimæ, per misericórdiam Dei, requiéscant in pace.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the souls of the faithful departed, through the mercy of God, rest in peace. \Rbar.~Amen.
}

%\flushpage

\heading{\section*{Marian Antiphon}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{alma-redemptoris-mater}
	
	\begin{versicles}
	\Vbar. Ángelus Dómini nuntiávit Maríæ.
	
	\Rbar. Et concépit de Spíritu Sancto.
	\end{versicles}
	
	\begin{collect}
	Grátiam tuam, qu\'{\ae}sumus, Dómine, méntibus nostris infúnde:
	ut qui, Ángelo nuntiánte, Christi Fílii tui incarnatiónem cognóvimus;
	per passiónem ejus et crucem, ad resurrectiónis gloriam perducámur.
	Per eum\-dem Christum Dóminum nostrum.
	\Amen
	\end{collect}
}{
	Mother of Christ, hear thou thy people's cry,
	Star of the deep and Portal of the sky!
	Mother of Him who thee from nothing made.
	Sinking we strive and call to thee for aid:
	oh, by what joy which Gabriel\linebreak brought to thee,
	thou Virgin first and last, let us thy mercy see. 
	
	\vfill
	
	The Angel of the Lord announced unto Mary.
	\Rbar.~And she conceived by the Holy Spirit.
	
	Let us pray.
	Pour forth, we beseech Thee, O Lord, Thy grace into our hearts:
	that as we have known the incarnation of Thy Son Jesus Christ by the message of an Angel,
	so too by His Cross and passion may we be brought to the glory of His resurrection.
	\Rbar.~Amen.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Divínum auxílium máneat semper nobíscum.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the divine assistance remain always with us.
	\Rbar.~Amen.
}

\end{latinenglishsection}

\pagebreak

\begin{latinenglishequalsection}

\spanning{\section*{Concluding Prayers}}

\rubric{All kneel and pray silently.}

\latinenglishequal{
	\begin{versicles}
	Sacrosánctæ et indivíduæ Trinitáti, Crucifíxi\linebreak Dómini nostri Jesu Christi humanitáti,
	beatíssimæ et gloriosíssimæ sempérque Vírginis Maríæ fecúndae integritáti,
	et ómnium Sanctórum universitáti sit sempitérna laus, honor, virtus et glória ab omni creatúra,
	nobísque remíssio ómnium peccatórum,
	per infiníta s\'{\ae}cula sæculórum. Amen.
	
	\Vbar. Beáta víscera Maríæ Vírginis, quæ portavérunt ætérni Patris Fílium.
	
	 \Rbar. Et beáta úbera quæ lactavérunt Christum\linebreak Dóminum. 
	 \end{versicles}
}{
	\begin{versicles}
	Everlasting praise, honor, power, and glory be given by all creatures to the most holy and undivided Trinity, to the Humanity of our crucified Lord Jesus Christ, to the fruitful purity of the most blessed and most glorious Mary ever Virgin, and to the company of all the Saints; and may we obtain the remission of all our sins through all eternity. Amen.
	
	\Vbar. Blessed is the womb of the Virgin Mary, that bore the Son of the eternal Father.
	
	 \Rbar. And blessed are the paps that gave suck to Christ our Lord.
	 \end{versicles}
}

\latinenglishequal{
	Pater Noster, qui es in cælis, sanctificétur nomen tuum.
	Advéniat regnum tuum.
	Fiat volúntas tua, sicut in cælo et in terra.
	Panem nostrum quoti\-diánum da nobis hódie,
	et dimítte nobis débita nostra, sicut et nos dimíttimus debitóribus nostris.
	Et ne nos indúcas in tentatiónem: sed líbera nos a malo. Amen.
}{
	Our Father, who art in heaven, hallowed be thy name.
	Thy kingdom come.
	Thy will be done, on earth as it is in heaven.
	Give us this day our daily bread, and forgive us our trespasses, as we forgive those who trespass against us.
	And lead us not into temptation: but deliver us from evil.
	Amen.
}
 
 \latinenglishequal{
	 Ave María, grátia plena, Dóminus tecum. Benedíc\-ta tu in muliéribus, et benedíctus fructus ventris tui, Jesus.
	 Sancta María, Mater Dei, ora pro nobis peccatóribus, nunc et in hora mortis nostræ. Amen.
}{
 	Hail Mary, full of grace, the Lord is with thee. Blessed art thou among women, and blessed is the fruit of thy womb, Jesus.
 	Holy Mary, Mother of God, pray for us sinners, now and at the hour of our death. Amen.
}
 
 \latinenglishequal{
	Credo in Deum, Patrem omnipoténtem, Creatórem cæli et terræ.
Et in Jesum Christum, Fílium ejus únicum, Dóminum nostrum: qui concéptus est de Spíritu Sancto, natus ex María Vírgine, passus sub Póntio Piláto, crucifíxus, mórtuus, et sepúltus: descéndit ad ínferos; tértia die resurréxit a mórtuis; ascéndit ad cælos; sedet ad déxteram Dei Patris omnipoténtis: inde ventúrus est judicáre vivos et mórtuos.
	Credo in Spíritum Sanctum, sanctam Ecclésiam cathólicam, Sanctórum communiónem, remissiónem peccatórum, carnis resurrectiónem, vitam ætérnam. Amen.
}{
	I believe in God,
	the Father Almighty,
	Creator of\linebreak heaven and earth,
	and in Jesus Christ, His only Son, our Lord,
	who was conceived by the Holy Spirit,
	born of the Virgin Mary,
	suffered under Pontius Pilate,
	was crucified, died and was buried;
	He descended into hell;
	on the third day He rose again from the dead;
	He ascended into heaven,
	and is seated at the right hand of God the Father Almighty;
	from thence He will come to judge the living and the dead.
	I believe in the Holy Spirit,
	the Holy Catholic Church,
	the communion of Saints,
	the forgiveness of sins,
	the resurrection of the body,
	and life everlasting.
	Amen.
}

\end{latinenglishequalsection}

\end{document}