\documentclass[11pt, twoside]{report}

\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage[bitstream-charter]{mathdesign}
\usepackage{bbding}
\usepackage{ragged2e}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{paracol}
\usepackage{mdframed}
\usepackage[margin=1in]{geometry}

\usepackage[autocompile]{gregoriotex}

\titleformat{\chapter}[block]{\huge\scshape\filcenter}{}{1em}{}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{}{1em}{}

\mdfsetup{skipabove=\topskip, skipbelow=\topskip}

\newcommand{\rubric}[1]{
	\switchcolumn[0] {
		\itshape
		#1
	}
}

\newcommand{\latinenglish}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape\small
		#2
	}
}

\newcommand{\latinenglishequal}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape
		#2
	}
}

\newenvironment{latinenglishsection}
	{\columnratio{.7, .3} \begin{paracol}{2}}
	{\end{paracol}}

\newenvironment{latinenglishequalsection}
	{\columnratio{.5, .5}\begin{paracol}{2}}
	{\end{paracol}}

\setlength{\columnseprule}{0.4pt}

\newcommand{\heading}[1]{
	\begin{leftcolumn}
		#1
	\end{leftcolumn}
}

\newcommand{\spanning}[1]{
	\switchcolumn*[#1]
}

\newenvironment{verses}[1]
	{\begin{flushleft} \begin{enumerate}[leftmargin=*] \setcounter{enumi}{#1}}
	{\end{enumerate} \end{flushleft}}

\newenvironment{versicles}{\par\leavevmode\parskip=0pt}{}

\newenvironment{collect}
{
	\leavevmode
	\parindent=1em
	\parskip=0pt
	\noindent Orémus.\par
}{}

\newenvironment{optionbox}
{
	\switchcolumn[0]
	\begin{mdframed}
%	\begin{minipage}{0.8\linewidth}
}{
%	\end{minipage}
	\end{mdframed}
}

\newcommand{\optionrule}{
	\begin{center}
	\rule{0.5\linewidth}{0.6pt}
	\end{center}
}

\newenvironment{optionruled}
{
	\optionrule
}
{
	\optionrule
}

% for use inside the collect environment
\newcommand{\Amen}{\par\noindent \Rbar. Amen.}

\begin{document}

\title{Vespers and Compline of Our Lady\\Throughout the Year}
\date{ }

\maketitle

\chapter*{Vespers}

\section*{Preparatory Prayers}

\textit{ All kneel and pray silently. As you say the prayer \textnormal{Aperi, Domine}, make the sign of the cross with your thumb first over your lips, and then over your heart.}

\begin{latinenglishequalsection}

\latinenglishequal{
	Áperi, \CrossMaltese\ Dómine, os meum ad benedicéndum\linebreak nomen sanctum tuum:
	\CrossMaltese\ munda quoque cor meum ab ómnibus vanis, pervérsis et aliénis cogitatiónibus;
	intelléctum illúmina, afféctum inflámma, ut digne, atténte ac devóte hoc Offícium beátæ Vírginis Maríæ recitáre váleam,
	et exaudíri mérear ante conspéctum divínæ Majestátis tuæ.
	Per Christum Dóminum nostrum. 
	Amen.
}{
	Open, + O Lord, my mouth to bless Thy holy Name; + cleanse also my heart from all vain, evil, and wandering thoughts; enlighten my understanding and kindle my affections; that I may worthily, attentively, and devoutly say this Office of the Blessed Virgin Mary, and so merit to be heard before the presence of Thy divine Majesty.  Through Christ our Lord.  Amen.
}

\latinenglishequal{
	Domine, in unióne illíus divínæ intentiónis, qua ipse in terris laudes Deo persolvísti, has tibi Horas persólvo.
}{
	O Lord, in union with that divine intention wherewith thou, whilst here on earth, didst render praises unto God, I desire to offer this my Office of prayer unto thee.
}

\latinenglishequal{
	Ave María, grátia plena, Dóminus tecum. Benedíc\-ta tu in muliéribus, et benedíctus fructus ventris tui, Jesus.
 Sancta María, Mater Dei, ora pro nobis peccatóribus, nunc et in hora mortis nostræ. Amen.
 }{
 	Hail Mary, full of grace, the Lord is with thee. Blessed art thou among women, and blessed is the fruit of thy womb, Jesus.
 Holy Mary, Mother of God, pray for us sinners, now and at the hour of our death. Amen.
 }
 
 \end{latinenglishequalsection}
 \pagebreak
 
 \begin{latinenglishsection}
 
\heading{\section*{Invitatory}}

\rubric{As the officiant intones the beginning of the office, all stand and make the sign of the cross.}

\latinenglish{
	\gregorioscore{deus-in-adjutorium}
}{
	O God, come to my assistance.
		
	O Lord, make haste to help me.
	
	Glory be to the Father, and to the Son, and to the Holy Spirit,
	as it was in the beginning, is now, and ever shall be, world without end. Amen.
	
	Alleluia.
}

\rubric{From Septuagesima until Easter, \textnormal{Alleluia} is replaced with}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{laus-tibi-domine}
}{
	Praise to thee, O Lord, King of everlasting glory.
}

\heading{\section*{Psalm 109}}

\rubric{The officiant intones the antiphon:}

\latinenglish{
	\textbf{Ant. 1.} Dum esset rex.
}{
	While the King.
}

\rubric{Then the cantor sings the first half of the first verse of the psalm, and the choir continues, alternating sides between each verse of the psalm.}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm1}
	
	\begin{verses}{1}
	
	\item Donec ponam ini\textbf{mí}cos \textbf{tu}os,~*\linebreak
	scabéllum pedum \textit{tu}\textbf{ó}rum.
	
	\item Virgam virtútis tuæ emíttet Dómi\textbf{nus} ex \textbf{Si}on:~*\linebreak
	domináre in médio inimicórum \textit{tu}\textbf{ó}rum.
	
	\item Tecum princípium in die virtútis tuæ in splendóri\textbf{bus} sanc\textbf{tó}rum:~*\linebreak
	ex útero ante lucíferum gé\textit{nu}\textbf{i} te.
	
	\item Jurávit Dóminus, et non p\oe ni\textbf{té}bit \textbf{e}um:~*\linebreak
	Tu es sacérdos in ætérnum secúndum órdinem \textit{Mel}\textbf{chí}sedech.
	
	\item Dóminus a \textbf{dex}tris \textbf{tu}is,~*\linebreak
	confrégit in die iræ su\textit{æ} \textbf{re}ges.
	
	\item Judicábit in natiónibus, im\textbf{plé}bit ru\textbf{í}nas:~*\linebreak
	conquassábit cápita in terra \textit{mul}\textbf{tó}rum.
	
	\item De torrénte in \textbf{vi}a \textbf{bi}bet:~*\linebreak
	\textit{(stand)} proptérea exaltá\textit{bit} \textbf{ca}put.
	
	\item \textit{(bow)} Glória \textbf{Pa}tri, et \textbf{Fí}\textbf{li}o,~*\linebreak
	et Spirítu\textit{i} \textbf{Sanc}to.
	
	\item \textit{(rise)} Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	\textit{(sit)} et in s\'{\ae}cula sæculó\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. The Lord said to my Lord:
	Sit thou at my right hand,
	
	2. Until I make thine enemies:
	thy footstool.
	
	3. The Lord shall send forth the rod of thy power from out of Sion:
	rule Thou in the midst of thine enemies.
	
	4. Thine shall be the dominion in the day of thy power, amid the brightness of the saints:
	from the womb, before the day-star, have I begotten thee.
	
	5. The Lord hath sworn, and will not repent:
	Thou art a priest forever according to the order of\linebreak Melchisedech.
	
	6. The Lord upon thy right hand:
	hath overthrown kings in the day of his wrath.
	
	7. He shall judge among the nations, he shall fulfil destructions:
	he shall smite in sunder the heads in the land of many.
	
	8. He shall drink of the brook in the way:
	therefore shall he lift up his head.
	
	Glory be.
}
	
\rubric{The same postures are observed at the end of every psalm.}
	
\rubric{All sing the full antiphon together:}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant1}
}{
	\textnormal{Ant.}
	While the King was reposing, my spikenard yielded the odor of sweetness.
}

\rubric{The remaining psalms with their antiphons are sung in the same way as the first.}

\heading{\section*{Psalm 112}}

\latinenglish{
	\textbf{Ant. 2.} Læva ejus.
}{
	His left hand.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm2}
	
	\begin{verses}{1}
	
	\item \textit{(bow head)} Sit nomen Dómini \textit{be}\textit{ne}\textbf{díc}tum,~*\linebreak
	ex hoc nunc, et \textit{us}\textit{que} \textit{in} \textbf{s\'{\ae}}culum.
	
	\item A solis ortu usque \textit{ad} \textit{oc}\textbf{cá}sum,~*\linebreak
	laudábi\textit{le} \textit{no}\textit{men} \textbf{Dó}mini.
	
	\item Excélsus super omnes \textit{gen}\textit{tes} \textbf{Dó}minus,~*\linebreak
	et super cælos \textit{gló}\textit{ri}\textit{a} \textbf{e}jus.
	
	\item Quis sicut Dóminus, Deus noster, qui in \textit{al}\textit{tis} \textbf{há}bitat,~*\linebreak
	et humília réspicit in cæ\textit{lo} \textit{et} \textit{in} \textbf{ter}ra?
	
	\item Súscitans a \textit{ter}\textit{ra} \textbf{ín}opem,~*\linebreak
	et de stércore \textit{é}\textit{ri}\textit{gens} \textbf{páu}perem:
	
	\item Ut cóllocet eum \textit{cum} \textit{prin}\textbf{cí}pibus,~*\linebreak
	cum princípibus \textit{pó}\textit{pu}\textit{li} \textbf{su}i.
	
	\item Qui habitáre facit stéri\textit{lem} \textit{in} \textbf{do}mo,~*\linebreak
	matrem fili\textit{ó}\textit{rum} \textit{læ}\textbf{tán}tem.
	
	\item Glória Pa\textit{tri}, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spi\textit{rí}\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textit{nunc}, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæ\textit{cu}\textit{ló}\textit{rum}. \textbf{A}men.

	\end{verses}
}{
	 1.  Praise the Lord, ye children:\linebreak praise ye the name of the Lord.
	 
	 2. Blessed be the name of the Lord, from henceforth now and for ever.
	 
	 3. From the rising of the sun unto the going down of the same, the name of the Lord is worthy of praise.
	 
	 4. The Lord is high above all nations; and his glory above the heavens.
	 
	 5. Who is as the Lord our God, who dwelleth on high: 
	 And looketh down on the low things in heaven and in earth?
	
	6. Raising up the needy from the earth, and lifting up the poor out of the dunghill:
	
	7. That he may place him with princes, with the princes of his people.
	
	8. Who maketh a barren woman to dwell in a house, the joyful\linebreak mother of children.
	
	Glory be.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant2}
}{
	\textnormal{Ant.}
	His left hand under my head, and his right hand shall embrace me.
}

\heading{\section*{Psalm 121}}

\latinenglish{
	\textbf{Ant. 3.} Nigra sum sed formósa.
}{
	I am black, but beautiful.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm3}

	\begin{verses}{1}
	
	\item Stantes erant \textbf{pe}des \textbf{nos}tri,~*\linebreak
	in átriis tuis, \textit{Je}\textbf{rú}salem.
	
	\item Jerúsalem, quæ ædifi\textbf{cá}tur ut \textbf{cí}\textbf{vi}tas:~*\linebreak
	cujus participátio ejus in \textit{id}\textbf{íp}sum.
	
	\item Illuc enim ascendérunt tribus, \textbf{tri}bus \textbf{Dó}\textbf{mi}ni:~*\linebreak
	testimónium Israël ad confiténdum nómi\textit{ni} \textbf{Dó}mini.
	
	\item Quia illic sedérunt sedes \textbf{in} ju\textbf{dí}\textbf{ci}o,~*\linebreak
	sedes super do\textit{mum} \textbf{Da}vid.
	
	\item Rogáte quæ ad pacem \textbf{sunt} Je\textbf{rú}\textbf{sa}lem:~*\linebreak
	et abundántia diligén\textit{ti}\textbf{bus} te:
	
	\item Fiat pax in vir\textbf{tú}te \textbf{tu}a:~*\linebreak
	et abundántia in túrri\textit{bus} \textbf{tu}is.
	
	\item Propter fratres meos, et \textbf{pró}ximos \textbf{me}os,~*\linebreak
	loquébar pa\textit{cem} \textbf{de} te:
	
	\item Propter domum Dómini, \textbf{De}i \textbf{nos}tri,~*\linebreak
	quæsívi bo\textit{na} \textbf{ti}bi.
	
	\item Glória \textbf{Pa}tri, et \textbf{Fí}\textbf{li}o,~*\linebreak
	et Spirítu\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæculó\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. I rejoiced at the things that were said to me: We shall go into the house of the Lord.
	
	2. Our feet were standing in thy courts, O Jerusalem.
	
	3. Jerusalem, which is built as a city, which is compact together.
	
	4. For thither did the tribes go up, the tribes of the Lord: the testimony of Israel, to praise the name of the Lord.
	
	5. Because their seats have sat in judgment, seats upon the house of David.

	6. Pray ye for the things that are for the peace of Jerusalem: and\linebreak abundance for them that love thee.
	
	7. Let peace be in thy strength: and abundance in thy towers.
	
	8. For the sake of my brethren, and of my neighbours, I spoke\linebreak peace of thee.
	
	9. Because of the house of the Lord our God, I have sought good things for thee.
	
	Glory be. 
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant3}
}{
	\textnormal{Ant.}
	I am black, but beautiful, O daughters of Jerusalem: therefore hath the king loved me, and brought me into his chamber.
}

\heading{\section*{Psalm 126}}

\latinenglish{
	\textbf{Ant. 4.} Jam hiems tránsiit.
}{
	Now is the winter past.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm4}
	
	\begin{verses}{1}
	
	\item Nisi Dóminus custodíerit civi\textbf{tá}tem,~*\linebreak
	frustra vígilat qui cus\textit{tó}\textit{dit} \textbf{e}am.
	
	\item Vanum est vobis ante lucem \textbf{súr}gere:~*\linebreak
	súrgite postquam sedéritis, qui manducátis pa\textit{nem} \textit{do}\textbf{ló}ris.
	
	\item Cum déderit diléctis suis \textbf{som}num:~*\linebreak
	ecce heréditas Dómini fílii: merces, \textit{fruc}\textit{tus} \textbf{ven}tris.
	
	\item Sicut sagíttæ in manu pot\textbf{én}tis:~*\linebreak
	ita fílii \textit{ex}\textit{cus}\textbf{só}rum.
	
	\item Beátus vir qui implévit desidérium suum ex \textbf{ip}sis:~*\linebreak
	non confundétur cum loquétur inimícis su\textit{is} \textit{in} \textbf{por}ta.
	
	\item Glória Patri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et nunc, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. Unless the Lord build the house, they labour in vain that build it.
	
	2. Unless the Lord keep the city, he watcheth in vain that keepeth it.
	
	3. It is vain for you to rise before light, rise ye after you have rested, you that eat the bread of sorrow.
	
	4. When he shall give sleep to his beloved: behold the inheritance of the Lord are children: the reward, the fruit of the womb.
	
	5. As arrows in the hand of the mighty, so the children of them that have been shaken.
	
	6. Blessed is the man that hath filled the desire with them;
	he shall not be confounded when he shall speak to his enemies in the gate. 
	
	Glory be.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant4}
}{
	\textnormal{Ant.}
	Now is the winter past, the rain is over and gone: arise, my beloved, and come.
}

\heading{\section*{Psalm 147}}

\latinenglish{
	\textbf{Ant. 5.} Speciósa facta es.
}{
	Thou art become beautiful.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm5}
	
	\begin{verses}{1}
	
	\item Quóniam confortávit seras portá\textit{rum} \textit{tu}\textbf{á}rum:~*\linebreak
	benedíxit fíli\textit{is} \textit{tu}\textit{is} \textbf{in} te.
	
	\item Qui pósuit fines \textit{tu}\textit{os} \textbf{pa}cem:~*\linebreak
	et ádipe frumén\textit{ti} \textit{sá}\textit{ti}\textbf{at} te.
	
	\item Qui emíttit elóquium \textit{su}\textit{um} \textbf{ter}ræ:~*\linebreak
	velóciter cur\textit{rit} \textit{ser}\textit{mo} \textbf{e}jus.
	
	\item Qui dat nivem \textit{sic}\textit{ut} \textbf{la}nam:~*\linebreak
	nébulam sicut \textit{cí}\textit{ne}\textit{rem} \textbf{spar}git.
	
	\item Mittit crystállum suam sic\textit{ut} \textit{buc}\textbf{cél}las:~*\linebreak
	ante fáciem frígoris ejus \textit{quis} \textit{sus}\textit{ti}\textbf{né}bit?
	
	\item Emíttet verbum suum, et liquefá\textit{ci}\textit{et} \textbf{e}a:~*\linebreak
	flabit spíritus ejus, \textit{et} \textit{flu}\textit{ent} \textbf{a}quæ.
	
	\item Qui annúntiat verbum \textit{su}\textit{um} \textbf{Ja}cob:~*\linebreak
	justítias, et judíci\textit{a} \textit{su}\textit{a} \textbf{Is}raël.
	
	\item Non fecit táliter omni \textit{na}\textit{ti}\textbf{ó}ni:~*\linebreak
	et judícia sua non mani\textit{fes}\textit{tá}\textit{vit} \textbf{e}is.
	
	\item Glória Pa\textit{tri}, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spi\textit{rí}\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textit{nunc}, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæ\textit{cu}\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	 1.  Praise the Lord, O Jerusalem: praise thy God, O Sion. 
	 
	 2. Because he hath strengthened the bolts of thy gates, he hath\linebreak blessed thy children within thee. 
	 
	 3. Who hath placed peace in thy borders: and filleth thee with the fat of corn.
	 
	 4. Who sendeth forth his speech to the earth: his word runneth swiftly.
	 
	 5. Who giveth snow like wool: scattereth mists like ashes.
	
	6. He sendeth his crystal like\linebreak morsels: who shall stand before the face of his cold?
	
	7. He shall send out his word, and shall melt them: his wind shall\linebreak blow, and the waters shall run.
	
	8. Who declareth his word to Jacob: his justices and his judgments to Israel.
	
	9. He hath not done in like manner to every nation: and his judgments he hath not made manifest to them.
	
	Glory be.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant5}
}{
	\textnormal{Ant.}
	Thou art become beautiful and sweet in thy delights, O holy Mother of God.
}

\heading{\section*{Little Chapter (Sirach 24:14)}}

\latinenglish{
	\textit{All stand.}
	
	Ab inítio et ante s\'{\ae}cula creáta sum,~\GreDagger\ 
	et usque ad futúrum s\'{\ae}culum non désinam,~* 
	et in habitatióne sancta coram ipso ministrávi.
	
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) <sp>R/</sp>. De(h)o(h) grá(f)ti(e)as.(e.f.) (::)
	}
}{
	From the beginning, and before the world, was I created,
	and unto the world to come I shall not cease to be,
	and in the holy dwelling place\linebreak I have ministered before him.\linebreak
	\Rbar.~Thanks be to God.
}

\heading{\section*{Hymn}}

\rubric{All kneel during the first verse of this hymn.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ave-maris-stella}
	
	\begin{verses}{1}
	
	\item \textit{(stand)} Sumens illud Ave, Gabriélis ore,\linebreak
	funda nos in pace, mutans Hevæ nomen.
	
	\item Solve vincla reis, profer lumen cæcis,\linebreak
	mala nostra pelle, bona cuncta posce.
	
	\item Monstra t\textit{e} esse matrem, sumat per te preces,\linebreak
	qui pro nobis natus, tulit esse tuus.
	
	\item Virgo singuláris, inter omnes mitis,\linebreak
	nos culpis solútos, mites fac et castos.
	
	\item Vitam præsta puram, iter para tutum,\linebreak
	ut vidéntes Jesum semper collætémur.
	
	\item \textit{(bow)} Sit laus Deo Patri, summo Christo decus,\linebreak
	Spirítui Sancto, tribus honor unus. Amen.
	
	\end{verses}
}{
	1.  Hail, O Star of the ocean,
	God's own Mother blest,
	ever sinless Virgin,
	gate of heav'nly rest. 
	
	2. Taking that sweet Ave,
	which from Gabriel came,
	peace confirm within us,
	changing Eve's name.
	
	3. Break the sinners' fetters,
	make our blindness day,
	chase all evils from us,
	for all blessings pray. 
	
	4. Show thyself a Mother,
	may the Word divine,
	born for us thine\linebreak Infant,
	hear our prayers through\linebreak thine.
	
	5. Virgin all excelling,
	mildest of the mild,
	free from guilt preserve us
	meek and undefiled.
	
	6. Keep our life all spotless,
	make our way secure,
	till we find in Jesus
	joy for evermore. 
	
	7. Praise to God the Father,
	honor to the Son,
	in the Holy Spirit,
	be the glory one. Amen. 
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{vers-diffusa-est}
}{
	Grace was poured forth on thy lips.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{vers-propterea-benedixit}
}{
	Therefore hath the Lord blessed thee for ever.
}

\heading{\section*{Canticle of Our Lady (Luke 1:46-55)}}

\rubric{Remain standing throughout the canticle.}

\latinenglish{
	\textbf{Ant. Magn.} Beáta Mater.
}{
	Blessed Mother.
}

\rubric{As the cantor intones the \textnormal{Magnificat}, make the sign of the cross.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{magnificat}
	
	\begin{verses}{2}
	
	\item Quia respéxit humilitátem \textit{an}\textit{cíl}\textit{læ} \textbf{su}æ:~*\linebreak
	ecce enim ex hoc beátam me dicent omnes genera\textit{ti}\textbf{ó}nes.
	
	\item Quia fecit mihi \textit{ma}\textit{gna} \textit{qui} \textbf{pot}ens est:~*\linebreak
	et sanctum no\textit{men} \textbf{e}jus.
	
	\item Et misericórdia ejus a progéni\textit{e} \textit{in} \textit{pro}\textbf{gé}nies~*\linebreak
	timénti\textit{bus} \textbf{e}um.
	
	\item Fecit poténtiam in \textit{brá}\textit{chi}\textit{o} \textbf{su}o:~*\linebreak
	dispérsit supérbos mente cor\textit{dis} \textbf{su}i.
	
	\item Depósuit pot\textit{én}\textit{tes} \textit{de} \textbf{se}de,~*\linebreak
	et exaltá\textit{vit} \textbf{hú}miles.
	
	\item Esuriéntes \textit{im}\textit{plé}\textit{vit} \textbf{bo}nis:~*\linebreak
	et dívites dimísit \textit{in}\textbf{á}nes.
	
	\item Suscépit Israël \textit{pú}\textit{e}\textit{rum} \textbf{su}um,~*\linebreak
	recordátus misericórdi\textit{æ} \textbf{su}æ.
	
	\item Sicut locútus est \textit{ad} \textit{pa}\textit{tres} \textbf{nos}tros,~*\linebreak
	Abraham et sémini ejus \textit{in} \textbf{s\'{\ae}}cula.
	
	\item  \textit{(bow)} Glória \textit{Pa}\textit{tri}, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spirítu\textit{i} \textbf{Sanc}to.
	
	\item  \textit{(rise)} Sicut erat in princípio, \textit{et} \textit{nunc}, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæculó\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. My soul doth magnify the Lord.
	
	2.  And my spirit hath rejoiced in God my Saviour.
	
	3. Because he hath regarded the humility of his handmaid;
	for\linebreak behold from henceforth all\linebreak generations shall call me blessed.
	
	4. Because he that is mighty, hath done great things to me; and holy is his name. 
	
	5. And his mercy is from generation unto generations, to them that fear him. 

	6. He hath shewed might in his arm: he hath scattered the proud in the conceit of their heart.
	
	7. He hath put down the mighty from their seat, and hath exalted the humble.
	
	8. He hath filled the hungry with good things; and the rich he hath sent empty away.
	
	9. He hath received Israel his servant, being mindful of his mercy:
	
	10. As he spoke to our fathers, to Abraham and to his seed for ever.
	
	Glory be.
}

\rubric{All sit.}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{antmagn}
}{
	\textnormal{Ant.}
	Blessed Mother and inviolate Virgin, glorious Queen of the world, intercede for us with the Lord.
}


\heading{\section*{Collect}}

\rubric{All stand.}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{	
	\begin{collect}
	Concéde nos fámulos tuos, qu\'{\ae}sumus, Dómine Deus, perpétua mentis et córporis sanitáte gaudére:
	et gloriósa beátæ Maríæ semper Vírginis intercessióne, a præsénti liberári tristítia, et ætérna pérfrui lætítia.
	Per Dóminum. \Amen
	\end{collect}
}{
	Let us pray.
	Grant, we beseech thee, O Lord God, that we, thy servants, may enjoy perpetual health, both of mind and body;
	and by the glorious intercession of blessed Mary ever Virgin,
	may be delivered from present sorrow, and attain unto\linebreak\ eternal joy.
	Through our Lord.\linebreak
	\Rbar.~Amen.
}

\heading{\section*{Commemorations (On Today's Handout)}}

\latinenglish{
	\textit{At this point there may follow one or more commemorations of the Sunday and of saints' feasts, found a separate handout.
	In the Roman Rite, Vespers is both the end of one liturgical day and the beginning of the next; as a result, there may be commemorations for feasts that fall either on today's date or tomorrow's.
	Each commemoration consists of an antiphon, a \textnormal{\Vbar} and \textnormal{\Rbar}, and an oration.
	Respond \textnormal{Amen} only at the end of the last commemoration.}
	
	\Rbar. Amen.
}{}

\heading{\section*{Suffrage of St. Joseph}}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant-joseph}
	
	\begin{versicles}
	\Vbar. Ecce homo sine queréla, verus Dei cultor.
	
	\Rbar. Ábstinens se ab omni ópere malo, et pérmanens in innocéntia sua.
	\end{versicles}
	
	\begin{collect}
	Sanctíssimæ Genetrícis tuæ Sponsi, qu\'{\ae}sumus, Dómine, méritis adjuvémur: ut, quod possibílitas nostra non óbtinet, ejus nobis intercessióne donétur.
	%(Qui vivis.)
	\end{collect}
}{
	A faithful and wise servant, whom the Lord hath set over His household.
	
	Behold a man without blame, a true worshipper of God.
	\Rbar.~Abstaining from every evil work, and abiding in his innocence.
	
	Let us pray.
	Assist us, O Lord, we beseech thee,
	by the merits of the spouse of thy most holy Mother,\linebreak
	that what of ourselves we are unable to obtain,
	may be granted to us by his intercession.
}

\heading{\section*{Suffrage of the Saints}}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant-all-saints}
	
	\begin{versicles}
	\Vbar. Lætámini in Dómino, et exsultáte, justi.
	
	\Rbar. Et gloriámini, omnes recti corde.
	\end{versicles}
	
	\begin{collect}
	Prótege, Dómine, pópulum tuum, et apostolórum tuórum Petri et Pauli, et aliórum apostolórum patrocínio confidéntem, perpétua defensióne consérva.
	
	Omnes Sancti tui, qu\'{\ae}sumus, Dómine, nos úbique ádjuvent;
	ut dum eórum mérita recólimus, patrocínia sentiámus:
	et pacem tuam nostris concéde tempóribus,
	et ab Ecclésia tua cunctam repélle nequítiam:
	iter, actus, et voluntátes nostras,
	et ómnium famulórum tuórum in salútis tuæ prosperitáte dispóne:
	benefactóribus nostris sempitérna bona\linebreak\ retríbue,
	et ómnibus fidélibus defúnctis réquiem ætérnam concéde.
	Per Dóminum nostrum Jesum Christum.
	\Amen
	\end{collect}
}{
	O all ye Saints of God, vouchsafe to intercede for our salvation, and that of all mankind.
	
	Rejoice in the Lord, and be glad, O ye just.
	\Rbar.~And glory, all ye that are right of heart.
	
	Let us pray. Protect thy people, O Lord, and preserve them by thy continual defence, who trust in the patronage of Peter and Paul, and all thy other apostles.
	
	Let all thy Saints, we beseech thee, O Lord, assist us everywhere;
	that, while we honor their merits, we\linebreak\ may experience their patronage:\linebreak
	grant us thy peace in our times, and repel all wickedness from thy Church:
	dispose our way, our acts, and wills, and those of all thy servants,
	in the good success of thy salvation:
	render to our benefactors everlasting blessings,
	and to all the faithful departed grant eternal rest.
	Through our Lord. \Rbar.~Amen.
}

\heading{\section*{Conclusion}}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{benedicamus-domino}
}{
	Let us bless the Lord.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{deo-gratias}
}{
	Thanks be to God.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Fidélium ánimæ, per misericórdiam Dei, requiéscant in pace.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the souls of the faithful departed, through the mercy of God, rest in peace. \Rbar.~Amen.
}

\end{latinenglishsection}

\textit{If Compline is not to be celebrated, finish with the Marian antiphon and concluding prayers.}

\chapter*{Before Compline}

\begin{latinenglishequalsection}

\spanning{\section*{Blessing}}

\rubric{The cantor asks for the blessing:}

\latinenglishequal{
	Jube, domne, benedícere.
}{
	Pray, Father, a blessing.
}

\latinenglishequal{
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}
	
	Jube, Dómine, benedícere.
	\end{optionbox}
}{
	\begin{optionruled}
	Pray, Lord, a blessing.
	\end{optionruled}
}

\rubric{The officiant replies,}

\latinenglishequal{
	\begin{versicles}
	Noctem quiétam et finem perféctum concédat nobis Dóminus omnípotens.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the almighty Lord grant unto us a peaceful night and a perfect end.
	\Rbar.~Amen.
}

\spanning{\section*{Short Lesson (1 Peter 5:8-9)}}

\latinenglishequal{
\begin{versicles}
	Fratres: Sóbrii estóte, et vigiláte: \GreDagger\ 
	quia adversárius vester diábolus, tamquam leo rúgiens, círcuit, qu\'{\ae}rens quem dévoret: *
	cui resístite fortes in fide.
	Tu autem Dómine miserére nobis.
	
	\Rbar. Deo grátias.
\end{versicles}
}{
	Brethren: be sober and watch,
	because your adversary, the devil, as a roaring lion, goeth about seeking whom he may devour,
	whom resist ye, strong in faith.
	But Thou, O Lord, have mercy on us.
	\Rbar.~Thanks be to God.
}

\spanning{\section*{Examination of Conscience}}

\latinenglishequal{
	\begin{versicles}
	\Vbar. Adjutórium \CrossMaltese\ nostrum in nómine Dómini.
	
	\Rbar. Qui fecit cælum et terram.
	\end{versicles}
}{
	Our help is in the name of the Lord.
	\Rbar.~Who made heaven and earth.
}

\rubric{All pray together silently:}

\latinenglishequal{
	Pater Noster, qui es in cælis, sanctificétur nomen tuum.
	Advéniat regnum tuum.
	Fiat volúntas tua, sicut in cælo et in terra.
	Panem nostrum quoti\-diánum da nobis hódie,
	et dimítte nobis débita nostra, sicut et nos dimíttimus debitóribus nostris.
	Et ne nos indúcas in tentatiónem: sed líbera nos a malo.
	Amen.
}{
	Our Father, who art in heaven, hallowed be thy name.
	Thy kingdom come.
	Thy will be done, on earth as it is in heaven.
	Give us this day our daily bread, and forgive us our trespasses, as we forgive those who trespass against us.
	And lead us not into temptation: but deliver us from evil.
	Amen.
}

\spanning{\section*{Confession (with priest)}}

\latinenglishequal{
	\rubric{The priest makes his confession: \textnormal{Confíteor Deo omnipoténti\dots oráre pro me ad Dóminum, Deum nostrum}.}
}{
	I confess to almighty God\dots
}

\rubric{All bow slightly towards the priest and respond:}

\latinenglishequal{
	Misereátur tui omnípotens Deus, et, dimíssis peccátis tuis, perdúcat te ad vitam ætérnam.
}{
	May almighty God have mercy on thee, forgive thee thy sins, and bring thee to life everlasting.
}

\rubric{And the priest replies \textnormal{Amen}.}

\rubric{Then all bow profoundly and make their confession:}

\latinenglishequal{
	Confíteor Deo omnipoténti, beátæ Maríæ semper Vírgini, beáto Michaéli Archángelo, beáto Joánni Baptístæ, sanctis Apóstolis Petro et Paulo,\linebreak\ ómnibus Sanctis, et tibi, pater:
	quia peccávi nimis cogitatióne, verbo et ópere:
	\textit{(strike breast)} mea\linebreak\ culpa, mea culpa, mea máxima culpa.
	 Ideo precor beátam Maríam semper Vírginem, beátum\linebreak\ Michaélem Archángelum,
	 beátum Joánnem Baptístam, sanctos Apóstolos Petrum et Paulum,
	 omnes Sanctos, et te, pater,
	 oráre pro me ad Dóminum, Deum nostrum.
}{
	I confess to almighty God, to Blessed Mary ever Virgin, to blessed Michael the Archangel, to blessed John the Baptist, to the holy Apostles Peter and Paul, to all the Saints, and to thee, Father, that I have sinned exceedingly in thought, word, and deed:
	through my fault, through my fault, through my most grievous fault.
	Therefore I pray Blessed Mary ever Virgin,\linebreak\ blessed Michael the Archangel, blessed John the Baptist, the holy Apostles Peter and Paul, all the saints, and thee, Father, to pray for me to the Lord our God.
}

\rubric{Remain bowed as the priest responds:}

\latinenglishequal{
	\begin{versicles}
	Misereátur vestri omnípotens Deus, et, dimíssis\linebreak\ peccátis vestris, perdúcat vos ad vitam ætérnam.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May almighty God have mercy on you, forgive you your sins, and bring you to life everlasting. 
	\Rbar.~Amen.
}

\rubric{Arise and make the sign of the cross as he says:}

\latinenglishequal{
	\begin{versicles}
	Indulgéntiam, \CrossMaltese\ absolutiónem, et remissiónem\linebreak\ peccatórum nostrórum
	tríbuat nobis omnípotens et miséricors Dóminus.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the almighty and merciful Lord grant us pardon, absolution, and remission of our sins.
	\Rbar.~Amen.
}

\spanning{\section*{Confession (without priest)}}


\rubric{All bow profoundly and say together:}
	
\latinenglishequal{	
	Confíteor Deo omnipoténti, beátæ Maríæ semper Vírgini, beáto Michaéli Archángelo, beáto Joánni Baptístæ, sanctis Apóstolis Petro et Paulo, et ómnibus Sanctis: quia peccávi nimis cogitatióne, verbo et ópere:
	\textit{(strike breast)} mea culpa, mea culpa, mea máxima culpa.
	Ideo precor beátam Maríam semper Vírginem, beátum Michaélem Archángelum,
	beátum Joánnem Baptístam, sanctos Apóstolos Petrum et Paulum,
	et omnes Sanctos, oráre pro me ad\linebreak\ Dóminum, Deum nostrum.
}{
	I confess to almighty God, to Blessed Mary ever Virgin, to blessed Michael the Archangel, to blessed John the Baptist, to the holy Apostles Peter and Paul, and to all the Saints, that I have sinned exceedingly in thought, word, and deed: through my fault, through my fault, through my most grievous fault.
	Therefore I pray Blessed Mary ever Virgin, blessed Michael the Archangel, blessed John the Baptist, the holy Apostles Peter and Paul, all the saints, and thee, Father, to pray for me to the Lord our God.
}

\latinenglishequal{
	Misereátur nostri omnípotens Deus, et, dimíssis\linebreak\ peccátis nostris, perdúcat nos ad vitam ætérnam. Amen.
}{
	May almighty God have mercy on us, forgive us our sins, and bring us to life everlasting. Amen.
}

\rubric{All arise and make the sign of the cross as they say:}

\latinenglishequal{
	Indulgéntiam, \CrossMaltese\ absolutiónem, et remissiónem\linebreak\ peccatórum nostrórum tríbuat nobis omnípotens et miséricors Dóminus. Amen.
}{
	May the almighty and merciful Lord grant us pardon, absolution, and remission of our sins. Amen.
}

\end{latinenglishequalsection}

\chapter*{Compline}

\begin{latinenglishsection}

\heading{\section*{Invitatory}}

\rubric{Cross yourself over your heart with your thumb as the officiant begins:}

\latinenglish{
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) <sp>V/</sp>. Con(h)vér(h)te(h) nos(h), <v>\CrossMaltese</v>() De(h)us(h), sa(h)lu(h)tá(g)ris(f) nos(h)ter.(h.) (::)
	}
}{
	Convert us, O God our Savior.
}

\latinenglish{
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) <sp>R/</sp>. Et(h) a(h)vér(h)te(h) i(h)ram(h) tu(h)am(g) a(f) no(h)bis.(h.) (::)
	}
}{
	And turn away Thine anger from us.
}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{deus-in-adjutorium-compl}
}{
	O God, come to my assistance.
		
	O Lord, make haste to help me.
	
	Glory be to the Father, and to the Son, and to the Holy Spirit,
	as it was in the beginning, is now, and ever shall be, world without end. Amen.
	
	Alleluia.
}
	

\rubric{From Septuagesima until Easter, \textnormal{Alleluia} is replaced with}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{laus-tibi-domine}
}{
	Praise to thee, O Lord, King of everlasting glory.
}

\rubric{The psalms of compline are all sung using the same tone, without antiphons.}

\heading{\section*{Psalm 128}}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm1-compl}
	
	\begin{verses}{1}
	
	\item Sæpe expugnavérunt me a juventú\textit{te} \textbf{me}a:~*\linebreak
	étenim non potu\textit{é}\textit{runt} \textbf{mi}hi.
	
	\item Supra dorsum meum fabricavérunt pec\textit{ca}\textbf{tó}res:~*\linebreak
	prolongavérunt iniqui\textit{tá}\textit{tem} \textbf{su}am.
	
	\item Dóminus justus concídit cervíces pec\textit{ca}\textbf{tó}rum:~*\linebreak
	confundántur et convertántur retrórsum omnes, qui o\textit{dé}\textit{runt} \textbf{Si}on.
	
	\item Fiant sicut f\oe num \textit{tec}\textbf{tó}rum:~*\linebreak
	quod priúsquam evellá\textit{tur} \textit{ex}\textbf{á}ruit:
	
	\item De quo non implévit manum suam \textit{qui} \textbf{me}tit:~*\linebreak
	et sinum suum qui maní\textit{pu}\textit{los} \textbf{cól}ligit.
	
	\item Et non dixérunt qui præteríbant: Benedíctio Dómi\textit{ni} \textbf{su}per vos:~*\linebreak
	benedíximus vobis in nó\textit{mi}\textit{ne} \textbf{Dó}mini.
	
	\item Glória Patri, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et nunc, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. Often have they fought against me from my youth, let Israel now say.
	
	2. Often have they fought against me from my youth: but they could not prevail over me.
	
	3. The wicked have wrought upon my back: they have lengthened\linebreak\ their iniquity.
	
	4. The Lord who is just will cut the necks of sinners: Let them all be confounded and turned back that hate Sion.

	5. Let them be as grass on the tops of houses: which withered before it be plucked up.
	
	6. Wherewith the mower filleth not his hand: nor he that gathereth sheaves his bosom.
	
	7. And they that have passed by have not said: The blessing of the Lord be upon you:
	we have blessed you in the name of the Lord. 
	
	Glory be.
}

\heading{\section*{Psalm 129}}

\latinenglish{
	\begin{verses}{0}
	
	\item De profúndis clamávi ad \textit{te}, \textbf{Dó}mine:~*\linebreak
	Dómine, exáudi \textit{vo}\textit{cem} \textbf{me}am:
	
	\item Fiant aures tuæ in\textit{ten}\textbf{dén}tes:~*\linebreak
	in vocem deprecati\textit{ó}\textit{nis} \textbf{me}æ.
	
	\item Si iniquitátes observáve\textit{ris}, \textbf{Dó}mine:~*\linebreak
	Dómine, quis \textit{sus}\textit{ti}\textbf{né}bit?
	
	\item Quia apud te propitiá\textit{ti}\textbf{o} est:~*\linebreak
	et propter legem tuam sustínu\textit{i} \textit{te}, \textbf{Dó}mine.
	
	\item Sustínuit ánima mea in ver\textit{bo} \textbf{e}jus:~*\linebreak
	sperávit ánima me\textit{a} \textit{in} \textbf{Dó}mino.
	
	\item A custódia matutína usque \textit{ad} \textbf{noc}tem:~*\linebreak
	speret Isra\textit{ël} \textit{in} \textbf{Dó}mino.
	
	\item Quia apud Dóminum mise\textit{ri}\textbf{cór}dia:~*\linebreak
	et copiósa apud e\textit{um} \textit{red}\textbf{émp}tio.
	
	\item Et ipse rédi\textit{met} \textbf{Is}raël:~*\linebreak
	ex ómnibus iniquitá\textit{ti}\textit{bus} \textbf{e}jus.
	
	\item Glória Patri, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et nunc, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
 	1. Out of the depths I have cried to thee, O Lord:
 	Lord, hear my voice.
 	
 	2. Let thine ears be attentive to the voice of my supplication. 
 	
 	3. If thou, O Lord, wilt mark iniquities: Lord, who shall stand it?
 	
 	4.For with thee there is merciful\linebreak\ forgiveness:
 	and by reason of thy law, I have waited for thee, O Lord.
 	
 	5. My soul hath relied on his word: my soul hath hoped in the Lord.

	6. From the morning watch even until night, let Israel hope in the Lord.
	
	7. Because with the Lord there is mercy: and with him plentiful redemption.
	
	8. And he shall redeem Israel from all his iniquities. 
	
	Glory.
}

\heading{\section*{Psalm 130}}

\latinenglish{
	\begin{verses}{0}
	
	\item Dómine, non est exaltátum \textit{cor} \textbf{me}um:~*\linebreak
	neque eláti sunt ó\textit{cu}\textit{li} \textbf{me}i.
	
	\item Neque ambulávi \textit{in} \textbf{ma}gnis:~*\linebreak
	neque in mirabí\textit{li}\textit{bus} \textbf{su}per me.
	
	\item Si non humíliter sen\textit{ti}\textbf{é}bam:~*\linebreak
	sed exaltávi á\textit{ni}\textit{mam} \textbf{me}am.
	
	\item Sicut ablactátus est super ma\textit{tre} \textbf{su}a:~*\linebreak
	ita retribútio in á\textit{ni}\textit{ma} \textbf{me}a.
	
	\item Speret Israël \textit{in} \textbf{Dó}mino:~*\linebreak
	ex hoc nunc et us\textit{que} \textit{in} \textbf{s\'{\ae}}culum.
	
	\item Glória Patri, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et nunc, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1.  Lord, my heart is not exalted: nor are my eyes lofty.
	
	2. Neither have I walked in great matters, nor in wonderful things above me.
	
	3.  If I was not humbly minded, but exalted my soul:
	
	4. As a child that is weaned is towards his mother, so reward in my soul.
	
	5. Let Israel hope in the Lord, from henceforth now and for ever. 
	
	Glory.
}

\heading{\section*{Hymn}}

\rubric{All stand.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{memento-salutis-auctor}
	
	\begin{verses}{1}
	
	\item María, Mater grátiæ, Mater misericórdiæ,\linebreak
	tu nos ab hoste prótege, et hora mortis súscipe.
	
	\item \textit{(bow)} Glória tibi, Dómine, qui natus es de Vírgine,\linebreak
	cum Patr\textit{e} et Sancto Spíritu, in sempitérna s\'{\ae}cula. Amen.
	
	\end{verses}
}{
	1.  Remember, O Creator Lord,
	that in the Virgin's sacred womb
	Thou wast conceived, and of her flesh\linebreak
	didst our mortality assume.
	
	2. Mother of grace, O Mary blest,
	to thee, sweet fount of love, we fly;
	shield us through life, and take us hence
	to thy dear bosom when we die.
	
	3. O Jesu! born of Mary bright!
	Immortal glory be to Thee;
	praise to the Father infinite,
	and Holy\linebreak\ Ghost eternally. Amen. 
}

\heading{\section*{Little Chapter (Sirach 24:24)}}

\latinenglish{
	Ego mater pulchræ dilectiónis,~\GreDagger\ et timóris, et agnitiónis,~* et sanctæ spei.
	
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) <sp>R/</sp>. De(h)o(h) grá(f)ti(e)as.(e.f.) (::)
	}
}{
	I am the mother of fair love, and of fear, and of knowledge, and of holy hope.
	\Rbar.~Thanks be to God.
}

\heading{\section*{Short Responsory}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ave-maria-rbr}
}{
	Hail Mary, full of grace: the Lord is with thee.
	\Rbar.~Hail Mary\dots.
	\Vbar.~Blessed art thou among women, and blessed is the fruit of thy womb.
	\Rbar.~The Lord is with thee.
	\Vbar.~Glory be to the Father, and to the Son, and to the Holy Spirit.
	\Rbar.~Hail Mary\dots.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Ora pro nobis, sancta Dei Génitrix.
	
	\Rbar. Ut digni efficiámur promissiónibus Christi.
	\end{versicles}
}{
	Pray for us, O holy Mother of God.
	\Rbar.~That we may be made worthy of the promises of Christ.
}

\heading{\section*{Canticle of Simeon (Luke 2:29-32)}}

\rubric{Remain standing throughout the canticle.}

\latinenglish{
	\textbf{Ant. Nunc.} Sub tuum præsídium confúgimus.
}{
	We fly to thy patronage.
}

\rubric{As the cantor intones the \textnormal{Nunc dimittis}, make the sign of the cross.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{nunc-dimittis}

	\begin{verses}{1}
	
	\item Quia vidérunt \textbf{ó}culi \textbf{me}i~*\linebreak
	salu\textbf{tá}re \textbf{tu}um,
	
	\item \textbf{Quod} pa\textbf{rás}ti~*\linebreak
	ante fáciem ómnium \textbf{po}pu\textbf{ló}rum,
	
	\item Lumen ad revelati\textbf{ó}nem \textbf{Gén}tium,~*\linebreak
	et glóriam plebis \textbf{tu}æ \textbf{Is}raël.
	
	\item \textit{(bow)} Glória \textbf{Pa}tri, et \textbf{Fí}lio,~*\linebreak
	et Spi\textbf{rí}tui \textbf{Sanc}to.
	
	\item  \textit{(rise)} Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textbf{ló}rum. \textbf{A}men.
	
	\end{verses}
}{
	1. Now thou dost dismiss thy servant, O Lord, according to thy word in peace.
	
	2. Because my eyes have seen thy salvation,

	3. Which thou hast prepared before the face of all peoples.
	
	4. A light to the revelation of the Gentiles, and the glory of thy people Israel. 
	
	Glory be.
}

\rubric{All sit.}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{ant-nunc-dimittis}
}{
	\textnormal{Ant.}
	We fly to thy patronage, O holy\linebreak\ Mother of God: despise not our petitions in our necessities,
	but deliver us always from all dangers, O glorious and blessed Virgin.
}

\heading{\section*{Collect and Conclusion}}

\rubric{All stand.}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{	
	\begin{collect}
	Beátæ et gloriósæ semper Vírginis Maríæ, qu\'{\ae}sumus, Dómine, intercéssio gloriósa nos prótegat: et ad vitam perdúcat ætérnam.
	Per Dominum nostrum Jesus Christum.
	\Amen
	\end{collect}
}{
	Let us pray. May the glorious intercession of the blessed and glorious Mary ever Virgin, protect us, we beseech Thee, O Lord, and bring us to life everlasting.
	Through our Lord.
	\Rbar.~Amen.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{
	\begin{versicles}
	\Vbar. Benedicámus Dómino.
	
	\Rbar. Deo grátias.
	\end{versicles}
}{
	Let us bless the Lord.
	\Rbar.~Thanks be to God.
}

\latinenglish{
	\begin{versicles}
	Benedícat et custódiat nos omnípotens et miséricors Dóminus, \CrossMaltese\ Pater, et Fílius, et Spíritus Sanctus.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the almighty and merciful Lord, Father, and Son, and Holy Spirit, bless us and preserve us.
	\Rbar.~Amen.
}

\flushpage

\heading{\section*{Marian Antiphon (before Easter)}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ave-regina-caelorum}
	
	\begin{versicles}
	\Vbar. Dignáre me laudáre te, Virgo sacráta.
	
	\Rbar. Da mihi virtútem contra hostes tuos.
	\end{versicles}
	
	\begin{collect}
	Concéde, miséricors Deus, fragilitáti nostræ præsídium:
	ut qui\linebreak sanctæ Dei Genitrícis memóriam ágimus;
	intercessiónis ejus auxílio, a nostris iniquitátibus resurgámus.
	Per eúmdem Christum Dóminum nostrum.
	\Amen
	\end{collect}
}{
	Hail, O Queen of Heaven enthroned.
	Hail, by angels mistress owned.\linebreak
	Root of Jesse, Gate of Morn
	whence the world's true light was born.
	
	Glorious Virgin, Joy to thee,
	loveliest whom in heaven they see;
	fairest thou, where all are fair,
	plead with Christ our souls to spare.
	
	\vfill
	
	Vouchsafe that I may praise thee, O sacred Virgin.
	\Rbar.~Give me stength against thine enemies.
	
	Grant, O merciful God, to our weak natures Thy protection,
	that we\linebreak\ who commemorate the holy Mother of God may,
	by the help of her intercession, arise from our iniquities.
	Through the same Christ our Lord.
	\Rbar.~Amen.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Divínum auxílium máneat semper nobíscum.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the divine assistance remain always with us.
	\Rbar.~Amen.
}

\flushpage

\heading{\section*{Marian Antiphon (after Pentecost)}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{salve-regina}
	
	\begin{versicles}
	\Vbar. Ora pro nobis, sancta Dei Génitrix.
	
	\Rbar. Ut digni efficiámur promissiónibus Christi.
	\end{versicles}
	
	\begin{collect}
	Omnípotens sempitérne Deus, qui gloriósæ Vírginis Matris Maríæ corpus et ánimam, ut dignum Fílii tui habitáculum éffici mererétur, Spíritu Sancto cooperánte, præparásti: da, ut cujus commemoratióne lætámur, ejus pia intercessióne ab instántibus malis et a morte perpétua liberémur.
	Per eúmdem Christum Dóminum nostrum.
	\Amen
	\end{collect}
}{
	Hail holy Queen, Mother of mercy, our life, our sweetness, and our\linebreak hope.
	To thee do we cry, poor banished children of Eve;
	to thee do we send up our sighs, mourning and weeping in this valley of tears.
	Turn then, O most gracious advocate,\linebreak thine eyes of mercy towards us,
	and after this our exile, show unto us the blessed fruit of thy womb, Jesus.
	O clement, O loving, O sweet virgin Mary.
	
	\vfill
	
	Pray for us, O holy Mother of God.
	\Rbar.~That we may be made worthy of the promises of Christ.
	
	Let us pray.
	O almighty everlasting God, who, by the cooperation of the Holy Ghost,
	didst prepare the body and soul of Mary, glorious Virgin and Mother,
	to become the worthy habitation of thy Son;
	grant that we may be delivered from\linebreak present evils and from everlasting death by her pious intercession,
	in whose commemoration we rejoice.
	Through the same Christ our Lord.
	\Rbar.~Amen.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Divínum auxílium máneat semper nobíscum.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the divine assistance remain always with us.
	\Rbar.~Amen.
}

\end{latinenglishsection}

\pagebreak

\begin{latinenglishequalsection}

\spanning{\section*{Concluding Prayers}}

\rubric{All kneel and pray silently.}

\latinenglishequal{
	\begin{versicles}
	Sacrosánctæ et indivíduæ Trinitáti, Crucifíxi\linebreak Dómini nostri Jesu Christi humanitáti,
	beatíssimæ et gloriosíssimæ sempérque Vírginis Maríæ fecúndae integritáti,
	et ómnium Sanctórum universitáti sit sempitérna laus, honor, virtus et glória ab omni creatúra,
	nobísque remíssio ómnium peccatórum,
	per infiníta s\'{\ae}cula sæculórum. Amen.
	
	\Vbar. Beáta víscera Maríæ Vírginis, quæ portavérunt ætérni Patris Fílium.
	
	 \Rbar. Et beáta úbera quæ lactavérunt Christum\linebreak Dóminum. 
	 \end{versicles}
}{
	\begin{versicles}
	Everlasting praise, honor, power, and glory be given by all creatures to the most holy and undivided Trinity, to the Humanity of our crucified Lord Jesus Christ, to the fruitful purity of the most blessed and most glorious Mary ever Virgin, and to the company of all the Saints; and may we obtain the remission of all our sins through all eternity. Amen.
	
	\Vbar. Blessed is the womb of the Virgin Mary, that bore the Son of the eternal Father.
	
	 \Rbar. And blessed are the paps that gave suck to Christ our Lord.
	 \end{versicles}
}

\latinenglishequal{
	Pater Noster, qui es in cælis, sanctificétur nomen tuum.
	Advéniat regnum tuum.
	Fiat volúntas tua, sicut in cælo et in terra.
	Panem nostrum quoti\-diánum da nobis hódie,
	et dimítte nobis débita nostra, sicut et nos dimíttimus debitóribus nostris.
	Et ne nos indúcas in tentatiónem: sed líbera nos a malo. Amen.
}{
	Our Father, who art in heaven, hallowed be thy name.
	Thy kingdom come.
	Thy will be done, on earth as it is in heaven.
	Give us this day our daily bread, and forgive us our trespasses, as we forgive those who trespass against us.
	And lead us not into temptation: but deliver us from evil.
	Amen.
}
 
 \latinenglishequal{
	 Ave María, grátia plena, Dóminus tecum. Benedíc\-ta tu in muliéribus, et benedíctus fructus ventris tui, Jesus.
	 Sancta María, Mater Dei, ora pro nobis peccatóribus, nunc et in hora mortis nostræ. Amen.
}{
 	Hail Mary, full of grace, the Lord is with thee. Blessed art thou among women, and blessed is the fruit of thy womb, Jesus.
 	Holy Mary, Mother of God, pray for us sinners, now and at the hour of our death. Amen.
}
 
 \latinenglishequal{
	Credo in Deum, Patrem omnipoténtem, Creatórem cæli et terræ.
Et in Jesum Christum, Fílium ejus únicum, Dóminum nostrum: qui concéptus est de Spíritu Sancto, natus ex María Vírgine, passus sub Póntio Piláto, crucifíxus, mórtuus, et sepúltus: descéndit ad ínferos; tértia die resurréxit a mórtuis; ascéndit ad cælos; sedet ad déxteram Dei Patris omnipoténtis: inde ventúrus est judicáre vivos et mórtuos.
	Credo in Spíritum Sanctum, sanctam Ecclésiam cathólicam, Sanctórum communiónem, remissiónem peccatórum, carnis resurrectiónem, vitam ætérnam. Amen.
}{
	I believe in God,
	the Father Almighty,
	Creator of\linebreak heaven and earth,
	and in Jesus Christ, His only Son, our Lord,
	who was conceived by the Holy Spirit,
	born of the Virgin Mary,
	suffered under Pontius Pilate,
	was crucified, died and was buried;
	He descended into hell;
	on the third day He rose again from the dead;
	He ascended into heaven,
	and is seated at the right hand of God the Father Almighty;
	from thence He will come to judge the living and the dead.
	I believe in the Holy Spirit,
	the Holy Catholic Church,
	the communion of Saints,
	the forgiveness of sins,
	the resurrection of the body,
	and life everlasting.
	Amen.
}

\end{latinenglishequalsection}

\end{document}