\documentclass[11pt, twoside]{report}

\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage[bitstream-charter]{mathdesign}
\usepackage{bbding}
\usepackage{ragged2e}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{paracol}
\usepackage{mdframed}
\usepackage[margin=1in]{geometry}

\usepackage[autocompile]{gregoriotex}

\titleformat{\chapter}[block]{\huge\scshape\filcenter}{}{1em}{}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{}{1em}{}

\mdfsetup{skipabove=\topskip, skipbelow=\topskip}

\newcommand{\rubric}[1]{
	\switchcolumn[0] {
		\itshape
		#1
	}
}

\newcommand{\latinenglish}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape\small
		#2
	}
}

\newcommand{\latinenglishequal}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape
		#2
	}
}

\newenvironment{latinenglishsection}
	{\columnratio{.7, .3} \begin{paracol}{2}}
	{\end{paracol}}

\newenvironment{latinenglishequalsection}
	{\columnratio{.5, .5}\begin{paracol}{2}}
	{\end{paracol}}

\setlength{\columnseprule}{0.4pt}

\newcommand{\heading}[1]{
	\begin{leftcolumn}
		#1
	\end{leftcolumn}
}

\newcommand{\spanning}[1]{
	\switchcolumn*[#1]
}

\newenvironment{verses}[1]
	{\begin{flushleft} \begin{enumerate}[leftmargin=*] \setcounter{enumi}{#1}}
	{\end{enumerate} \end{flushleft}}

\newenvironment{versicles}{\par\leavevmode\parskip=0pt}{}

\newenvironment{collect}
{
	\leavevmode
	\parindent=1em
	\parskip=0pt
	\noindent Orémus.\par
}{}

\newenvironment{optionbox}
{
	\switchcolumn[0]
	\begin{mdframed}
%	\begin{minipage}{0.8\linewidth}
}{
%	\end{minipage}
	\end{mdframed}
}

\newcommand{\optionrule}{
	\begin{center}
	\rule{0.5\linewidth}{0.6pt}
	\end{center}
}

\newenvironment{optionruled}
{
	\optionrule
}
{
	\optionrule
}

% for use inside the collect environment
\newcommand{\Amen}{\par\noindent \Rbar. Amen.}

\begin{document}

\chapter*{Vespers of Our Lady in Christmastide}

\textit{This form of the Office of Vespers is celebrated for the first time on Christmas Eve,
and for the last time on February 2nd, that is, the Feast of the Purification. 
However, if the date of the Purification is transferred, the change in the Office is not transferred with it.}

\section*{Prayers Before the Office}

\textit{ All kneel and pray silently. As you say the prayer \textnormal{Aperi, Domine}, make the sign of the cross with your thumb first over your lips, and then over your heart.}

\begin{latinenglishequalsection}

\latinenglishequal{
	Aperi, + Dómine, os meum ad benedicéndum\linebreak nomen sanctum tuum:
	+ munda quoque cor meum ab ómnibus vanis, pervérsis et aliénis cogitatió-\linebreak nibus;
	intelléctum illúmina, afféctum inflámma, ut digne, atténte ac devóte hoc Offícium beátæ Vírginis Maríæ recitáre váleam,
	et exaudíri mérear ante conspéctum divínæ Majestátis tuæ.
	Per Christum Dóminum nostrum. 
	Amen.
}{
	Open, + O Lord, my mouth to bless Thy holy Name; + cleanse also my heart from all vain, evil, and wandering thoughts; enlighten my understanding and kindle my affections; that I may worthily, attentively, and devoutly say this Office of the Blessed Virgin Mary, and so merit to be heard before the presence of Thy divine Majesty.  Through Christ our Lord.  Amen.
}

\latinenglishequal{
	Domine, in unióne illíus divínæ intentiónis, qua ipse in terris laudes Deo persolvísti, hanc tibi Horam persólvo.
}{
	O Lord, in union with that divine intention wherewith Thou, whilst here on earth, didst render praises unto God, I desire to offer this Hour of prayer unto Thee.
}

\latinenglishequal{
	Ave María, grátia plena, Dóminus tecum. Benedíc\-ta tu in muliéribus, et benedíctus fructus ventris tui, Jesus.
 Sancta María, Mater Dei, ora pro nobis peccatóribus, nunc et in hora mortis nostræ. Amen.
 }{
 	Hail Mary, full of grace, the Lord is with thee. Blessed art thou among women, and blessed is the fruit of thy womb, Jesus.
 Holy Mary, Mother of God, pray for us sinners, now and at the hour of our death. Amen.
 }
 
 \end{latinenglishequalsection}
 \pagebreak
 
 \begin{latinenglishsection}
 
\heading{\section*{Invitatory}}

\rubric{As the officiant intones the beginning of the office, all stand and make the sign of the cross.}

\latinenglish{
	\gregorioscore{../per-annum/deus-in-adjutorium}
}{
	O God, come to my assistance.
		
	O Lord, make haste to help me.
	
	Glory be to the Father, and to the Son, and to the Holy Spirit,
	as it was in the beginning, is now, and ever shall be, world without end. Amen.
	
	Alleluia.
}

\rubric{After Septuagesima, \textnormal{Alleluia} is replaced with}

\latinenglish{
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) Laus(h) ti(h)bi(h) Dó(h)mi(h)ne(h'_) Rex(h) æ(h)tér(h)næ(i) gló-(h')ri-(h)æ.(g.) (::)
	}
}{
	Praise to thee, O Lord, King of everlasting glory.
}

\heading{\section*{Psalm 109}}

\rubric{The officiant intones the antiphon:}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{intonation1-xmas}
}{
	O marvelous interchange!
}

\rubric{The cantor sings the first verse of the psalm, and all sit halfway through the verse, at the asterisk.}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm1-xmas}
	
	\begin{verses}{1}
	
	\item Donec ponam ini\textbf{mí}cos \textbf{tu}os,~*\linebreak
	scabéllum pe\textit{dum} \textit{tu}\textbf{ó}rum.

	\item Virgam virtútis tuæ emíttet Dómi\textbf{nus} ex \textbf{Si}on:~*\linebreak
	domináre in médio inimicó\textit{rum} \textit{tu}\textbf{ó}rum.
	
	\pagebreak
	
	\item Tecum princípium in die virtútis tuæ in splendóri\textbf{bus} sanc\textbf{tó}rum:~*\linebreak
	ex útero ante lucíferum \textit{gé}\textit{nu}\textbf{i} te.
	
	\item Jurávit Dóminus, et non p\oe ni\textbf{té}bit \textbf{e}um:~*\linebreak
	Tu es sacérdos in ætérnum secúndum órdi\textit{nem} \textit{Mel}\textbf{chí}sedech.
	
	\item Dóminus a \textbf{dex}tris \textbf{tu}is,~*\linebreak
	confrégit in die iræ \textit{su}\textit{æ} \textbf{re}ges.
	
	\item Judicábit in natiónibus, im\textbf{plé}bit ru\textbf{í}nas:~*\linebreak
	conquassábit cápita in ter\textit{ra} \textit{mul}\textbf{tó}rum.
	
	\item De torrénte in \textbf{vi}a \textbf{bi}bet:~*\linebreak
	proptérea exal\textit{tá}\textit{bit} \textbf{ca}put.
	
	\item \textit{(bow)} Glória \textbf{Pa}tri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item \textit{(rise)} Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in  s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
	
	\rubric{The bow at the \textnormal{Gloria Patri} is observed at the end of every psalm.}

	\gresetinitiallines{1}
	\gregorioscore{ant1-xmas}
}{
	1. The Lord said to my Lord:
	Sit thou at my right hand,
	
	2. Until I make thine enemies:
	thy footstool.
	
	3. The Lord shall send forth the rod of thy power from out of Sion:
	rule Thou in the midst of thine enemies.
	
	\pagebreak
	
	4. Thine shall be the dominion in the day of thy power, amid the brightness of the saints:
	from the womb, before the day-star, have I begotten thee.
	
	5. The Lord hath sworn, and will not repent:
	Thou art a priest forever according to the order of\linebreak Melchisedech.
	
	6. The Lord upon thy right hand:
	hath overthrown kings in the day of his wrath.
	
	7. He shall judge among the nations, he shall fulfil destructions:
	he shall smite in sunder the heads in the land of many.
	
	8. He shall drink of the brook in the way:
	therefore shall he lift up his head.
	
	Glory be.
	
	O marvelous interchange! The Creator of mankind, taking a body with a living soul, vouchsafed to be born of a Virgin,
	and becoming man without man's concurrence, bestowed upon us his Deity.
}


%\latinenglish{
%	\gresetinitiallines{1}
%	\gregorioscore{ant1-xmas}
%}{
%	O marvelous interchange! The Creator of mankind, taking a body with a living soul, vouchsafed to be born of a Virgin,
%	and becoming man without man's concurrence, bestowed upon us his Deity.
%}


\heading{\section*{Psalm 112}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{intonation2-xmas}
}{
	When Thou wast born.
}

\flushpage

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm2-xmas}
	
	\begin{verses}{1}
	
	\item \textit{(bow head)} Sit nomen Dómini \textbf{be}ne\textbf{díc}tum,~*\linebreak
	ex hoc nunc, et us\textit{que} \textit{in} \textbf{s\'{\ae}}culum.
	
	\item A solis ortu usque \textbf{ad} oc\textbf{cá}sum,~*\linebreak
	laudábile \textit{no}\textit{men} \textbf{Dó}mini.
	
	\item Excélsus super omnes \textbf{gen}tes \textbf{Dó}\textbf{mi}nus,~*\linebreak
	et super cælos gló\textit{ri}\textit{a} \textbf{e}jus.
	
	\item Quis sicut Dóminus, Deus noster, qui in \textbf{al}tis \textbf{há}\textbf{bi}tat,~*\linebreak
	et humília réspicit in cælo \textit{et} \textit{in} \textbf{ter}ra?
	
	\item Súscitans a \textbf{ter}ra \textbf{ín}\textbf{o}pem,~*\linebreak
	et de stércore é\textit{ri}\textit{gens} \textbf{páu}perem:
	
	\item Ut cóllocet eum \textbf{cum} prin\textbf{cí}\textbf{pi}bus,~*\linebreak
	cum princípibus pó\textit{pu}\textit{li} \textbf{su}i.
	
	\item Qui habitáre facit stéri\textbf{lem} in \textbf{do}mo,~*\linebreak
	matrem filió\textit{rum} \textit{læ}\textbf{tán}tem.
	
	\item Glória \textbf{Pa}tri, et \textbf{Fí}\textbf{li}o,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.

	\end{verses}
}{
	 1.  Praise the Lord, ye children:\linebreak praise ye the name of the Lord.
	 
	 2. Blessed be the name of the Lord, from henceforth now and for ever.
	 
	 3. From the rising of the sun unto the going down of the same, the name of the Lord is worthy of praise.
	 
	 4. The Lord is high above all nations; and his glory above the heavens.
	 
	 5. Who is as the Lord our God, who dwelleth on high: 
	 And looketh down on the low things in heaven and in earth?
	
	6. Raising up the needy from the earth, and lifting up the poor out of the dunghill:
	
	7. That he may place him with princes, with the princes of his people.
	
	8. Who maketh a barren woman to dwell in a house, the joyful\linebreak mother of children.
	
	Glory be.
}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant2-xmas}
}{
	When Thou wast born of a Virgin, after an ineffable manner, then were the Scriptures fulfilled:
	Thou didst come down like rain upon the fleece, that Thou mightest save mankind: we praise Thee, O our God.
}

\heading{\section*{Psalm 121}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{intonation3-xmas}
}{
	In the bush which Moses saw.
}

\flushpage

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm3-xmas}

	\begin{verses}{1}
	
	\item Stantes erant \textit{pe}\textit{des} \textbf{nos}tri,~*\linebreak
	in átriis \textit{tu}\textit{is}, \textit{Je}\textbf{rú}\textbf{sa}lem.

	\item Jerúsalem, quæ ædificá\textit{tur} \textit{ut} \textbf{cí}vitas:~*\linebreak
	cujus participátio e\textit{jus} \textit{in} \textit{id}\textbf{íp}sum.
	
	\item Illuc enim ascendérunt tribus, \textit{tri}\textit{bus} \textbf{Dó}mini:~*\linebreak
	testimónium Israël ad confiténdum \textit{nó}\textit{mi}\textit{ni} \textbf{Dó}\textbf{mi}ni.
	
	\item Quia illic sedérunt sedes \textit{in} \textit{ju}\textbf{dí}cio,~*\linebreak
	sedes su\textit{per} \textit{do}\textit{mum} \textbf{Da}vid.
	
	\item Rogáte quæ ad pacem \textit{sunt} \textit{Je}\textbf{rú}salem:~*\linebreak
	et abundántia di\textit{li}\textit{gén}\textit{ti}\textbf{bus} te:
	
	\item Fiat pax in vir\textit{tú}\textit{te} \textbf{tu}a:~*\linebreak
	et abundántia in \textit{túr}\textit{ri}\textit{bus} \textbf{tu}is.
	
	\item Propter fratres meos, et pró\textit{xi}\textit{mos} \textbf{me}os,~*\linebreak
	loqué\textit{bar} \textit{pa}\textit{cem} \textbf{de} te:
	
	\item Propter domum Dómini, \textit{De}\textit{i} \textbf{nos}tri,~*\linebreak
	quæsí\textit{vi} \textit{bo}\textit{na} \textbf{ti}bi.
	
	\item Glória Pa\textit{tri}, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spi\textit{rí}\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textit{nunc}, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæ\textit{cu}\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. I rejoiced at the things that were said to me: We shall go into the house of the Lord.
	
	2. Our feet were standing in thy courts, O Jerusalem.
	
	3. Jerusalem, which is built as a city, which is compact together.
	
	4. For thither did the tribes go up, the tribes of the Lord: the testimony of Israel, to praise the name of the Lord.
	
	5. Because their seats have sat in judgment, seats upon the house of David.

	6. Pray ye for the things that are for the peace of Jerusalem: and\linebreak abundance for them that love thee.
	
	7. Let peace be in thy strength: and abundance in thy towers.
	
	8. For the sake of my brethren, and of my neighbours, I spoke\linebreak peace of thee.
	
	9. Because of the house of the Lord our God, I have sought good things for thee.
	
	Glory be. 
}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant3-xmas}
}{
	In the bush which Moses saw unconsumed, we acknowledge thy admirable virginity preserved:
	intercede for us, O Mather of God.
}

\flushpage

\heading{\section*{Psalm 126}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{intonation4-xmas}
}{
	The root hath budded.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm4-xmas}
	
	\begin{verses}{1}

	\item Nisi Dóminus custodíerit \textbf{ci}vi\textbf{tá}tem,~*\linebreak
	frustra vígilat qui cus\textit{tó}\textit{dit} \textbf{e}am.

	\item Vanum est vobis ante \textbf{lu}cem \textbf{súr}gere:~*\linebreak
	súrgite postquam sedéritis, qui manducátis pa\textit{nem} \textit{do}\textbf{ló}ris.
	
	\item Cum déderit diléctis \textbf{su}is \textbf{som}num:~*\linebreak
	ecce heréditas Dómini fílii: merces, \textit{fruc}\textit{tus} \textbf{ven}tris.
	
	\item Sicut sagíttæ in \textbf{ma}nu pot\textbf{én}tis:~*\linebreak
	ita fílii \textit{ex}\textit{cus}\textbf{só}rum.
	
	\item Beátus vir qui implévit desidérium \textbf{su}um ex \textbf{ip}sis:~*\linebreak
	non confundétur cum loquétur inimícis su\textit{is} \textit{in} \textbf{por}ta.
	
	\item Glória \textbf{Pa}tri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. Unless the Lord build the house, they labour in vain that build it.
	
	2. Unless the Lord keep the city, he watcheth in vain that keepeth it.
	
	3. It is vain for you to rise before light, rise ye after you have rested, you that eat the bread of sorrow.
	
	4. When he shall give sleep to his beloved: behold the inheritance of the Lord are children: the reward, the fruit of the womb.
	
	5. As arrows in the hand of the mighty, so the children of them that have been shaken.
	
	6. Blessed is the man that hath filled the desire with them;
	he shall not be confounded when he shall speak to his enemies in the gate. 
	
	Glory be.
}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant4-xmas}
}{
	The root of Jesse hath budded, a star hath arisen out of Jacob; a virgin hath borne the Savior: we praise Thee, O our God.
}

\heading{\section*{Psalm 147}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{intonation5-xmas}
}{
	Behold, Mary.
}

\flushpage

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{psalm5-xmas}
	
	\begin{verses}{1}
	
	\item Quóniam confortávit seras portárum tu\textbf{á}rum:~*\linebreak
	benedíxit fíliis tu\textit{is} \textbf{in} te.

	\item Qui pósuit fines tuos \textbf{pa}cem:~*\linebreak
	et ádipe fruménti sá\textit{ti}\textbf{at} te.
	
	\item Qui emíttit elóquium suum \textbf{ter}ræ:~*\linebreak
	velóciter currit ser\textit{mo} \textbf{e}jus.
	
	\item Qui dat nivem sicut \textbf{la}nam:~*\linebreak
	nébulam sicut cíne\textit{rem} \textbf{spar}git.
	
	\item Mittit crystállum suam sicut buc\textbf{cél}las:~*\linebreak
	ante fáciem frígoris ejus quis sus\textit{ti}\textbf{né}bit?
	
	\item Emíttet verbum suum, et liquefáciet \textbf{e}a:~*\linebreak
	flabit spíritus ejus, et flu\textit{ent} \textbf{a}quæ.
	
	\item Qui annúntiat verbum suum \textbf{Ja}cob:~*\linebreak
	justítias, et judícia su\textit{a} \textbf{Is}raël.
	
	\item Non fecit táliter omni nati\textbf{ó}ni:~*\linebreak
	et judícia sua non manifestá\textit{vit} \textbf{e}is.
	
	\item Glória Patri, et \textbf{Fí}lio,~*\linebreak
	et Spirítu\textit{i} \textbf{Sanc}to.
	
	\item Sicut erat in princípio, et nunc, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæculó\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	 1.  Praise the Lord, O Jerusalem: praise thy God, O Sion. 
	 
	 2. Because he hath strengthened the bolts of thy gates, he hath\linebreak blessed thy children within thee. 
	 
	 3. Who hath placed peace in thy borders: and filleth thee with the fat of corn.
	 
	 4. Who sendeth forth his speech to the earth: his word runneth swiftly.
	 
	 5. Who giveth snow like wool: scattereth mists like ashes.
	
	6. He sendeth his crystal like\linebreak morsels: who shall stand before the face of his cold?
	
	7. He shall send out his word, and shall melt them: his wind shall\linebreak blow, and the waters shall run.
	
	8. Who declareth his word to Jacob: his justices and his judgments to Israel.
	
	9. He hath not done in like manner to every nation: and his judgments he hath not made manifest to them.
	
	Glory be.
}

\rubric{After Septuagesima, the antiphon is sung without \textnormal{Alleluia}.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant5-xmas}
}{
	Behold, Mary hath borne us the Savior, whom John beholding, exclaimed, saying:
	Behold the Lamb of God, behold Him who taketh away the sins of the world, (alleluia).
}

\heading{\section*{Little Chapter (Sirach 24:14)}}

\latinenglish{	
	\gresetinitiallines{1}
	\gabcsnippet{
	(c3) 	AB(h) i(h)ní(h)ti(h)o(h) et(h) an(h)te(h) s<sp>'ae</sp>(h)cu(h)la(h) cre(h)á(h)ta(f) sum,(f)
	et(h) us(h)que(h) ad(h) fu(h)tú(h)rum(h) s<sp>'ae</sp>(h)cu(h)lum(g) non(f) dé(h)si(h)nam,(h)
	et(h) in(h) ha(h)bi(h)ta(h)ti(h)ó(h)ne(h) sanc(h)ta(h) co(h)ram(h) ip(h)so(h) mi(h)nis(h)trá(f)vi.(ef..) (::)
	<sp>R/</sp>. De(h)o(h) grá(f)ti(e)as.(ef..) (::)
	}
}{
	From the beginning, and before the world, was I created,
	and unto the world to come I shall not cease to be,
	and in the holy dwelling place\linebreak I have ministered before him.\linebreak
	\Rbar.~Thanks be to God.
}

\heading{\section*{Hymn}}

\rubric{All kneel during the first verse of this hymn.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{../per-annum/ave-maris-stella}
}{
	1.  Hail, O Star of the ocean,
	God's own Mother blest,
	ever sinless Virgin,
	gate of heav'nly rest. 
	
	2. Taking that sweet Ave,
	which from Gabriel came,
	peace confirm within us,
	changing Eve's name.
	
	3. Break the sinners' fetters,
	make our blindness day,
	chase all evils from us,
	for all blessings pray. 
	
	4. Show thyself a Mother,
	may the Word divine,
	born for us thine\linebreak Infant,
	hear our prayers through\linebreak thine.
	
	5. Virgin all excelling,
	mildest of the mild,
	free from guilt preserve us
	meek and undefiled.
	
	\pagebreak
	
	6. Keep our life all spotless,
	make our way secure,
	till we find in Jesus
	joy for evermore. 
	
	7. Praise to God the Father,
	honor to the Son,
	in the Holy Spirit,
	be the glory one. Amen. 
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{../per-annum/vers-diffusa-est}
}{
	Grace was poured forth on thy lips.
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{../per-annum/vers-propterea-benedixit}
}{
	Therefore hath the Lord blessed thee for ever.
}

\heading{\section*{Canticle of Our Lady (Luke 1:46-55)}}

\rubric{Remain standing throughout the canticle.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{intonationmagn-xmas}
}{
	A great mystery.
}

\rubric{As the cantor intones the \textnormal{Magnificat}, make the sign of the cross.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{magnificat-xmas}
	
	\begin{verses}{2}
	
	\item Quia respéxit humilitátem \textit{an}\textit{cíl}\textit{læ} \textbf{su}æ:~*\linebreak
	ecce enim ex hoc beátam me dicent omnes genera\textit{ti}\textbf{ó}nes.

	\item Quia fecit mihi \textit{ma}\textit{gna} \textit{qui} \textbf{pot}ens est:~*\linebreak
	et sanctum no\textit{men} \textbf{e}jus.
	
	\item Et misericórdia ejus a progéni\textit{e} \textit{in} \textit{pro}\textbf{gé}nies~*\linebreak
	timénti\textit{bus} \textbf{e}um.
	
	\item Fecit poténtiam in \textit{brá}\textit{chi}\textit{o} \textbf{su}o:~*\linebreak
	dispérsit supérbos mente cor\textit{dis} \textbf{su}i.
	
	\item Depósuit pot\textit{én}\textit{tes} \textit{de} \textbf{se}de,~*\linebreak
	et exaltá\textit{vit} \textbf{hú}miles.
	
	\item Esuriéntes \textit{im}\textit{plé}\textit{vit} \textbf{bo}nis:~*\linebreak
	et dívites dimísit \textit{in}\textbf{á}nes.
	
	\item Suscépit Israël \textit{pú}\textit{e}\textit{rum} \textbf{su}um,~*\linebreak
	recordátus misericórdi\textit{æ} \textbf{su}æ.
	
	\item Sicut locútus est \textit{ad} \textit{pa}\textit{tres} \textbf{nos}tros,~*\linebreak
	Abraham et sémini ejus \textit{in} \textbf{s\'{\ae}}cula.
	
	\item \textit{(bow)} Glória \textit{Pa}\textit{tri}, \textit{et} \textbf{Fí}lio,~*\linebreak
	et Spirítu\textit{i} \textbf{Sanc}to.
	
	\item \textit{(rise)} Sicut erat in princípio, \textit{et} \textit{nunc}, \textit{et} \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæculó\textit{rum}. \textbf{A}men.
	
	\end{verses}
}{
	1. My soul doth magnify the Lord.
	
	2.  And my spirit hath rejoiced in God my Saviour.
	
	3. Because he hath regarded the humility of his handmaid;
	for\linebreak behold from henceforth all\linebreak generations shall call me blessed.
	
	4. Because he that is mighty, hath done great things to me; and holy is his name. 
	
	5. And his mercy is from generation unto generations, to them that fear him. 

	6. He hath shewed might in his arm: he hath scattered the proud in the conceit of their heart.
	
	7. He hath put down the mighty from their seat, and hath exalted the humble.
	
	8. He hath filled the hungry with good things; and the rich he hath sent empty away.
	
	9. He hath received Israel his servant, being mindful of his mercy:
	
	10. As he spoke to our fathers, to Abraham and to his seed for ever.
	
	Glory be.
}

\rubric{All sit.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{antmagn-xmas}
}{
	A great mystery of inheritance: the womb of one that knew not man hath become the temple of God;
	taking flesh of her he was not defiled: all nations shall come, saying, Glory be to Thee, O Lord.
}


\heading{\section*{Collect}}

\rubric{All stand.}

\latinenglish{
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) Ký(h)ri(h)e(h) e(h)lé(h)i(h)son.(h.) <sp>R/</sp>.(::) Chris(h)te(h) e(h)lé(h)i(h)son.(h.) (,) Ký(h)ri(h)e(h) e(f)lé(g)i(h)son.(h.) (::)
	}
}{
	Lord have mercy.
	\Rbar.~Christ have mercy.
	Lord have mercy.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\pagebreak
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\pagebreak
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{	
	\begin{collect}
	Deus, qui salútis ætérnæ, beátæ Maríæ virginitáte fecúnda, humáno géneri pr\'{\ae}mia præstitísti:~\GreDagger\
	tríbue, qu\'{\ae}sumus; ut ipsam pro nobis intercédere sentiámus,~*
	per quam merúimus auctórem vitæ suscípere,
	Dóminum nostrum Jesum Christum Fílium tuum.
	\Amen
	\end{collect}
}{
	Let us pray.
	O God, who, by the fruitful virginity of the blessed Mary, hast given to mankind the rewards of eternal  salvation;
	grant, we beseech thee, that we may experience her intercession for us, through whom we have received the author of life,
	Thy Son Jesus Christ, our Lord.
	\Rbar.~Amen.
}


\heading{\section*{Suffrage of St. Joseph}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{../per-annum/ant-joseph}
}{
	Behold, a faithful and wise servant, whom the Lord hath set over His household.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Glória et divítiæ in domo ejus.
	
	\Rbar. Et justítia ejus manet in s\'{\ae}culum s\'{\ae}culi.
	\end{versicles}
	
	\begin{collect}
	Sanctíssimæ Genetrícis tuæ Sponsi, qu\'{\ae}sumus, Dómine, méritis adjuvémur:~\GreDagger\
	ut, quod possibílitas nostra non óbtinet,~*
	ejus nobis intercessióne donétur.
	%(Qui vivis.)
	\end{collect}
	
	\rubric{The conclusion of the prayer \textnormal{(Qui vivis)} is omitted, and the response \textnormal{Amen} is not made.}
}{
	Glory and wealth shall be in his house.
	\Rbar.~And his justice remaineth for ever and ever. 
	
	Let us pray.
	Assist us, O Lord, we beseech thee,
	by the merits of the spouse of thy most holy Mother,\linebreak
	that what of ourselves we are unable to obtain,
	may be granted to us by his intercession.
}

\heading{\section*{Suffrage of the Saints}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{../per-annum/ant-all-saints}
	
	\begin{versicles}
	\Vbar. Lætámini in Dómino, et exsultáte, justi.
	
	\Rbar. Et gloriámini, omnes recti corde.
	\end{versicles}
	
	\pagebreak
	
	\begin{collect}
	Prótege, Dómine, pópulum tuum,~*
	et apostolórum tuórum Petri et Pauli, et aliórum apostolórum patrocínio confidéntem, perpétua defensióne consérva.
	
	Omnes Sancti tui, qu\'{\ae}sumus, Dómine, nos úbique ádjuvent;
	ut dum eórum mérita recólimus, patrocínia sentiámus:~*
	et pacem tuam nostris concéde tempóribus,
	et ab Ecclésia tua cunctam repélle nequítiam:~**
	iter, actus, et voluntátes nostras,
	et ómnium famulórum tuórum in salútis tuæ prosperitáte dispóne:~*
	benefactóribus nostris sempitérna bona\linebreak\ retríbue,
	et ómnibus fidélibus defúnctis réquiem ætérnam concéde.
	
	Per Dóminum nostrum Jesum Christum, Filium tuum,~\GreDagger\
	qui tecum vivit et regnat in unitate Spiritus Sancti, Deus,~*
	per omnia s\'{\ae}cula sæculórum.
	\Amen
	\end{collect}
}{
	O all ye Saints of God, vouchsafe to intercede for our salvation, and that of all mankind.
	
	Rejoice in the Lord, and be glad, O ye just.
	\Rbar.~And glory, all ye that are right of heart.
	
	\pagebreak
	
	Let us pray. Protect thy people, O Lord, and preserve them by thy continual defence, who trust in the patronage of Peter and Paul, and all thy other apostles.
	
	Let all thy Saints, we beseech thee, O Lord, assist us everywhere;
	that, while we honor their merits, we\linebreak\ may experience their patronage:\linebreak
	grant us thy peace in our times, and repel all wickedness from thy Church:
	dispose our way, our acts, and wills, and those of all thy servants,
	in the good success of thy salvation:
	render to our benefactors everlasting blessings,
	and to all the faithful departed grant eternal rest.
	Through our Lord. \Rbar.~Amen.
}

%% force heading to come after the English translation of the long prayer
\latinenglish{
	\heading{\section*{Conclusion}}
}{}

\latinenglish{
	\begin{versicles}
	\Vbar. Dóminus vobíscum.

	\Rbar. Et cum spíritu tuo.
	\end{versicles}
	
	\begin{optionbox}
	\rubric{If the officiant is not a priest:}

	\begin{versicles}
	\Vbar. Dómine, exáudi oratiónem meam.

	\Rbar. Et clamor meus ad te véniat.
	\end{versicles}
	\end{optionbox}
}{
	The Lord be with you.
	\Rbar.~And with thy spirit.
	
	\begin{optionruled}
	Lord, hear my prayer.
	\Rbar.~And let my cry come unto thee.
	\end{optionruled}
}

\latinenglish{
	\gresetinitiallines{0}
	\gregorioscore{../per-annum/benedicamus-domino}
}{
	Let us bless the Lord. \Vbar.~Thanks be to God.
}

\rubric{In a low voice:}

\latinenglish{
	\begin{versicles}
	\Vbar. Fidélium ánimæ, per misericórdiam Dei, requiéscant in pace.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the souls of the faithful departed, through the mercy of God, rest in peace. \Rbar.~Amen.
}

\flushpage

\heading{\section*{Marian Antiphon}}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{../advent/alma-redemptoris-mater}
	
	\begin{versicles}
	\Vbar. Post partum, Virgo, invioláta permansísti.
	
	\Rbar. Dei Génitrix, intercéde pro nobis.
	\end{versicles}
	
	\begin{collect}
	Deus, qui salútis ætérnæ, beátæ Maríæ virginitáte fecúnda, humáno géneri pr\'{\ae}mia præstitísti:~\GreDagger\
	tríbue, qu\'{\ae}sumus; ut ipsam pro nobis intercédere sentiámus,~*
	per quam merúimus auctórem vitæ suscípere,
	Dóminum nostrum Jesum Christum Fílium tuum.
	\Amen
	\end{collect}
}{
	Mother of Christ, hear thou thy people's cry,
	Star of the deep and Portal of the sky!
	Mother of Him who thee from nothing made.
	Sinking we strive and call to thee for aid:
	oh, by what joy which Gabriel brought to thee,
	thou Virgin first and last, let us thy mercy see. 
	
	\vfill
	
	After child-birth thou didst remain a pure virgin.
	\Rbar.~Intercede for us, O Mother of God.
	
	Let us pray.
	O God, who, by the fruitful virginity of the blessed Mary, hast given to mankind the rewards of eternal  salvation;
	grant, we beseech thee, that we may experience her intercession for us, through whom we have received the author of life,
	Thy Son Jesus Christ, our Lord.
	\Rbar.~Amen.
}

\rubric{In a low voice:}

\latinenglish{
	\begin{versicles}
	\Vbar. Divínum auxílium máneat semper nobíscum.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the divine assistance remain always with us.
	\Rbar.~Amen.
}

\end{latinenglishsection}

\pagebreak

\begin{latinenglishequalsection}

\spanning{\section*{Prayers After the Office}}

\rubric{All kneel and pray silently.}
⎄
\latinenglishequal{
	Sacrosánctæ et indivíduæ Trinitáti, Crucifíxi\linebreak Dómini nostri Jesu Christi humanitáti,
	beatíssimæ et gloriosíssimæ sempérque Vírginis Maríæ fecúndae integritáti,
	et ómnium Sanctórum universitáti sit sempitérna laus, honor, virtus et glória ab omni creatúra,
	nobísque remíssio ómnium peccatórum,
	per infiníta s\'{\ae}cula sæculórum. Amen.
}{
	Everlasting praise, honor, power, and glory be given by all creatures to the most holy and undivided Trinity, to the Humanity of our crucified Lord Jesus Christ, to the fruitful purity of the most blessed and most glorious Mary ever Virgin, and to the company of all the Saints; and may we obtain the remission of all our sins through all eternity. Amen.
}

\latinenglishequal{
	\begin{versicles}
	\Vbar. Beáta víscera Maríæ Vírginis, quæ portavérunt ætérni Patris Fílium.
	
	 \Rbar. Et beáta úbera quæ lactavérunt Christum\linebreak Dóminum. 
	 \end{versicles}
}{
	\begin{versicles}	
	\Vbar. Blessed is the womb of the Virgin Mary, that bore the Son of the eternal Father.
	
	 \Rbar. And blessed are the paps that gave suck to Christ our Lord.
	 \end{versicles}
}

\latinenglishequal{
	Pater Noster, qui es in cælis, sanctificétur nomen tuum.
	Advéniat regnum tuum.
	Fiat volúntas tua, sicut in cælo et in terra.
	Panem nostrum quoti\-diánum da nobis hódie,
	et dimítte nobis débita nostra, sicut et nos dimíttimus debitóribus nostris.
	Et ne nos indúcas in tentatiónem: sed líbera nos a malo. Amen.
}{
	Our Father, who art in heaven, hallowed be thy name.
	Thy kingdom come.
	Thy will be done, on earth as it is in heaven.
	Give us this day our daily bread, and forgive us our trespasses, as we forgive those who trespass against us.
	And lead us not into temptation: but deliver us from evil.
	Amen.
}

\latinenglishequal{
	Ave María, grátia plena, Dóminus tecum. Benedíc\-ta tu in muliéribus, et benedíctus fructus ventris tui, Jesus.
 Sancta María, Mater Dei, ora pro nobis peccatóribus, nunc et in hora mortis nostræ. Amen.
 }{
 	Hail Mary, full of grace, the Lord is with thee. Blessed art thou among women, and blessed is the fruit of thy womb, Jesus.
 Holy Mary, Mother of God, pray for us sinners, now and at the hour of our death. Amen.
 }

\end{latinenglishequalsection}

\end{document}