\documentclass[11pt, twoside]{report}

\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage[bitstream-charter]{mathdesign}
\usepackage{bbding}
\usepackage{ragged2e}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{paracol}
\usepackage{mdframed}
\usepackage[margin=1in]{geometry}

\usepackage[autocompile]{gregoriotex}

\titleformat{\chapter}[block]{\huge\scshape\filcenter}{}{1em}{}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{}{1em}{}

\mdfsetup{skipabove=\topskip, skipbelow=\topskip}

\newcommand{\rubric}[1]{
	\switchcolumn[0] {
		\itshape
		#1
	}
}

\newcommand{\latinenglish}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape\small
		#2
	}
}

\newcommand{\latinenglishequal}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape
		#2
	}
}

\newenvironment{latinenglishsection}
	{\columnratio{.7, .3} \begin{paracol}{2}}
	{\end{paracol}}

\newenvironment{latinenglishequalsection}
	{\columnratio{.5, .5}\begin{paracol}{2}}
	{\end{paracol}}

\setlength{\columnseprule}{0.4pt}

\newcommand{\heading}[1]{
	\begin{leftcolumn}
		#1
	\end{leftcolumn}
}

\newcommand{\spanning}[1]{
	\switchcolumn*[#1]
}

\newenvironment{verses}[1]
	{\begin{flushleft} \begin{enumerate}[leftmargin=*] \setcounter{enumi}{#1}}
	{\end{enumerate} \end{flushleft}}

\newenvironment{versicles}{\par\leavevmode\parskip=0pt}{}

\newenvironment{collect}
{
	\leavevmode
	\parindent=1em
	\parskip=0pt
	\noindent Orémus.\par
}{}

\newenvironment{collect-no-oremus}
{
	\leavevmode
	\parindent=1em
	\parskip=0pt
}{}

\newenvironment{optionbox}
{
	\switchcolumn[0]
	\begin{mdframed}
%	\begin{minipage}{0.8\linewidth}
}{
%	\end{minipage}
	\end{mdframed}
}

\newcommand{\optionrule}{
	\begin{center}
	\rule{0.5\linewidth}{0.6pt}
	\end{center}
}

\newenvironment{optionruled}
{
	\optionrule
}
{
	\optionrule
}

% for use inside the collect environment
\newcommand{\Amen}{\par\noindent \Rbar. Amen.}

\begin{document}

\chapter*{Vespers of Our Lady in Eastertide}
\setcounter{page}{1}

\textit{This Office is celebrated for the first time on Easter Sunday, and for the last time on the Friday in the octave of Pentecost.
However, on the Feast of the Annunciation, which may fall in this period, the Office is that of Advent.
}

\textit{All is said as in the Office throughout the year, except for that which is given below, namely, the antiphon at the \textnormal{Magnificat} and the Marian anthem.}

\section*{Canticle of Our Lady (Luke 1:46-55)}

\begin{latinenglishsection}

\rubric{Remain standing throughout the canticle.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{intonation-regina-caeli}
}{
	O Queen of Heaven.
}

\rubric{As the cantor intones the \textnormal{Magnificat}, make the sign of the cross.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{magnificat-eastertide}
	
	\begin{verses}{2}
	
	\item Quia respéxit humilitátem an\textbf{cíl}læ \textbf{su}æ:~*\linebreak
	ecce enim ex hoc beátam me dicent omnes gene\textit{ra}\textit{ti}\textbf{ó}nes.
	
	\item Quia fecit mihi \textbf{ma}gna qui \textbf{pot}ens est:~*\linebreak
	et sanctum \textit{no}\textit{men} \textbf{e}jus.
	
	\item Et misericórdia ejus a progénie \textbf{in} pro\textbf{gé}nies~*\linebreak
	timén\textit{ti}\textit{bus} \textbf{e}um.
	
	\item Fecit poténtiam in \textbf{brá}chio \textbf{su}o:~*\linebreak
	dispérsit supérbos mente \textit{cor}\textit{dis} \textbf{su}i.
	
	\item Depósuit pot\textbf{én}tes de \textbf{se}de,~*\linebreak
	et exal\textit{tá}\textit{vit} \textbf{hú}\textbf{mi}les.
	
	\item Esuriéntes im\textbf{plé}vit \textbf{bo}nis:~*\linebreak
	et dívites dimí\textit{sit} \textit{in}\textbf{á}nes.
	
	\item Suscépit Israël \textbf{pú}erum \textbf{su}um,~*\linebreak
	recordátus misericór\textit{di}\textit{æ} \textbf{su}æ.
	
	\item Sicut locútus est ad \textbf{pa}tres \textbf{nos}tros,~*\linebreak
	Abraham et sémini e\textit{jus} \textit{in} \textbf{sǽ}\textbf{cu}la.
	
	\item  \textit{(bow)} Glória \textbf{Pa}tri, et \textbf{Fí}lio,~*\linebreak
	et Spirí\textit{tu}\textit{i} \textbf{Sanc}to.
	
	\item \textit{(rise)} Sicut erat in princípio, et \textbf{nunc}, et \textbf{sem}per,~*\linebreak
	et in s\'{\ae}cula sæcu\textit{ló}\textit{rum}. \textbf{A}men.
	\end{verses}
}{
	1. My soul doth magnify the Lord.
	
	2.  And my spirit hath rejoiced in God my Saviour.
	
	3. Because he hath regarded the humility of his handmaid;
	for\linebreak behold from henceforth all\linebreak generations shall call me blessed.
	
	4. Because he that is mighty, hath done great things to me; and holy is his name. 
	
	5. And his mercy is from generation unto generations, to them that fear him. 

	6. He hath shewed might in his arm: he hath scattered the proud in the conceit of their heart.
	
	7. He hath put down the mighty from their seat, and hath exalted the humble.
	
	\pagebreak
	
	8. He hath filled the hungry with good things; and the rich he hath sent empty away.
	
	9. He hath received Israel his servant, being mindful of his mercy:
	
	10. As he spoke to our fathers, to Abraham and to his seed for ever.
	
	Glory be.
}

\rubric{All sit.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant-regina-caeli}
}{
    	Joy to thee, O Queen of Heaven, alleluia!
    	He whom Thou wast meet to bear; alleluia.
   	As He promised, hath arisen; alleluia.
    	Pour for us to God thy prayer, alleluia.
}

\flushpage

\heading{\section*{Marian Anthem}}

\rubric{Remain standing on all days of Eastertide.}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{regina-caeli-anthem}
	
	\begin{versicles}
	\Vbar. Gaude et lætáre Virgo María, allelúia.
	
	\Rbar. Quia surréxit Dóminus vere, allelúia.
	\end{versicles}
	
	\begin{collect}
	Deus, qui per resurrectiónem Fílii tui Dómini nostri Jesu Christi\linebreak mundum lætificáre dignátus es:~\GreDagger\
	præsta, qu\'{\ae}sumus; ut per ejus Genitrícem Vírginem Maríam,~* perpétuæ capiámus gáudia vitæ.
	Per eúmdum Christum Dóminum nostrum.
	\Amen
	\end{collect}
}{
    	Joy to thee, O Queen of Heaven, alleluia!
    	He whom Thou wast meet to bear; alleluia.
   	As He promised, hath arisen; alleluia.
    	Pour for us to God thy prayer, alleluia.
	
	\vfill
	
	Rejoice and be glad, O Virgin Mary, alleluia.
	\Rbar.~For the Lord hath risen indeed: alleluia.
	
	O God, who didst vouchsafe to give joy to the world through the resurrection of thy Son our Lord Jesus Christ;
	grant, we beseech thee, that, through his Mother, the Virgin Mary, we may obtain the joys of everlasting life.
	Through the\linebreak same Christ our Lord.
	\Rbar.~Amen.
}

\latinenglish{
	\begin{versicles}
	\Vbar. Divínum auxílium máneat semper nobíscum.
	
	\Rbar. Amen.
	\end{versicles}
}{
	May the divine assistance remain always with us.
	\Rbar.~Amen.
}

\end{latinenglishsection}

\end{document}