\documentclass[11pt]{report}

\usepackage{fontspec}
\usepackage[utf8]{inputenc}
\usepackage[bitstream-charter]{mathdesign}
\usepackage{bbding}
\usepackage{ragged2e}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{paracol}
\usepackage{mdframed}
\usepackage[margin=1in]{geometry}

\usepackage[autocompile]{gregoriotex}

\titleformat{\chapter}[block]{\huge\scshape\filcenter}{}{1em}{}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{}{1em}{}

\mdfsetup{skipabove=\topskip, skipbelow=\topskip}

\newcommand{\rubric}[1]{
	\switchcolumn[0] {
		\itshape
		#1
	}
}

\newcommand{\latinenglish}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape\small
		#2
	}
}

\newcommand{\latinenglishequal}[2]{
	\switchcolumn[0]* {
		#1
	}
	\switchcolumn[1] {
		\itshape
		#2
	}
}

\newenvironment{latinenglishsection}
	{\columnratio{.7, .3} \begin{paracol}{2}}
	{\end{paracol}}

\newenvironment{latinenglishequalsection}
	{\columnratio{.5, .5}\begin{paracol}{2}}
	{\end{paracol}}

\setlength{\columnseprule}{0.4pt}

\newcommand{\heading}[1]{
	\begin{leftcolumn}
		#1
	\end{leftcolumn}
}

\newcommand{\spanning}[1]{
	\switchcolumn*[#1]
}

\newenvironment{verses}[1]
	{\begin{flushleft} \begin{enumerate}[leftmargin=*] \setcounter{enumi}{#1}}
	{\end{enumerate} \end{flushleft}}

\newenvironment{versicles}{\par\leavevmode\parskip=0pt}{}

\newenvironment{collect}
{
	\leavevmode
	\parindent=1em
	\parskip=0pt
	\noindent Orémus.\par
}{}

\newenvironment{optionbox}
{
	\switchcolumn[0]
	\begin{mdframed}
%	\begin{minipage}{0.8\linewidth}
}{
%	\end{minipage}
	\end{mdframed}
}

\newcommand{\optionrule}{
	\begin{center}
	\rule{0.5\linewidth}{0.6pt}
	\end{center}
}

\newenvironment{optionruled}
{
	\optionrule
}
{
	\optionrule
}

% for use inside the collect environment
\newcommand{\Amen}{\par\noindent \Rbar. Amen.}

\begin{document}

\chapter*{The Annunciation of Our Lord}

\textit{On the Feast of the Annunciation,  the Little Office of Advent is celebrated.
When the feast falls in Eastertide, the antiphons are to be used in the form given on this sheet.
The antiphons of the psalms and canticles should be doubled, that is, sung in full both before and after the psalm.}
 
 \begin{latinenglishsection}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant1-annunc-easter}
}{
	The angel Gabriel was sent to Mary, a virgin espoused to Joseph, alleluia.
}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant2-annunc-easter}
}{
	Hail Mary, full of grace, the Lord is with thee: blessed art thou among women, alleluia
}
 
 
\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant3-annunc-easter}
}{
	Fear not, Mary, thou hast found grace with the Lord: behold, thou shalt conceive, and bear a son, alleluia.
}


\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant4-annunc-easter}
}{
	The Lord God shall give unto him the throne of David his father: and he shall reign forever, alleluia.
}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{ant5-annunc-easter}
}{
	Behold the handmaid of the Lord: be it done unto me according to thy word, alleluia.
}

\latinenglish{
	\gresetinitiallines{1}
	\gregorioscore{antcant-annunc-easter}
}{
	The Holy Spirit shall come upon thee, Mary: fear not, thou shalt bear in thy womb the Son of God, alleluia.
}

\latinenglish{
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) <sp>V/</sp>. Dif(h)fú(h)sa(h) est(h) grá(h)ti(h)a(h) in(h) lá(h)bi(h)is(h) tu(h)is,(h)
	al(h)le(h)lú(h)ia.(g'_/hvGF'E/fgf.) (::)
	}
}{
	Grace was poured forth on thy lips, alleluia.
}

\latinenglish{
	\gresetinitiallines{0}
	\gabcsnippet{
	(c3) <sp>R/</sp>.  Prop(h)té(h)re(h)a(h) be(h)ne(h)dí(h)xit(h) te(h) De(h)us(h) in(h) æ(h)tér(h)num,(h) 
	al(h)le(h)lú(h)ia.(g'_/hvGF'E/fgf.) (::)
	}

}{
	Therefore hath the Lord blessed thee for ever, alleluia.
}


\end{latinenglishsection}

\end{document}